# Import
***

## Create Data

### Data Structures

#### Atomic Vector

`base`

* `c()`: Combine values into a vector or list.
* `factor()`: Create a factor variable.
* `seq()`: Generate a sequence.
* `vector()`: Create a vector.
    + This function is preferable to `c()` when creating an empty vector, according to the DataCamp article ["Five Tips to Improve Your R Code"](https://www.datacamp.com/community/tutorials/five-tips-r-code-improve).

#### Matrix

`base`

* `matrix()`: Create a matrix.

#### Array

`base`

* `array()`: Create an array.

#### List

`base`

* `c()`: Combine values into a vector or list.
* `list()`: Create a list.

`tibble`

* `tibble()`: Create a data frame or list.

#### Data Frame

`base`

* `data.frame()`: Create a data frame.
    + Prefer `tibble::tibble()` over `base::data.frame()`.
* `list()`: Create a list.

`tibble`

* `tibble()`: Create a data frame or list.

#### References

* ["Vectors"](https://r4ds.had.co.nz/vectors.html) (Grolemund & Wickham, [*R for Data Science*](https://r4ds.had.co.nz/))
* ["Data structures"](http://adv-r.had.co.nz/Data-structures.html) (Wickham, [*Advanced R*](http://adv-r.had.co.nz/))

### Other

`stats`

* `rnorm()`: Generate a random normal distribution.

***
## Import Data from a Local Drive

`base`

* `attach()`: Attach a set of R objects to the search path.
    + Allows objects in the database to be accessed by giving their names (e.g., `height` rather than `women$height`).
* `file.choose()`: Choose a file interactively.
    + Use as `file = file.choose()` in `read.table()` and similar functions.
* `load()`: Reload datasets saved with `save()`.
* `readRDS()`: Restore an R object written with `saveRDS()`.

`data.table`

* `fread()`: Similar to `read.table()`, but faster and more convenient for large data sets.

`foreign`

* `read.spss()`: Read an SPSS data file.

`haven`

* `read_sas()`: Read and write SAS files.

`readr`

* `read_delim()`: Read a delimited (including csv and txv) file.
    + Child functions: `read_csv()`, `read_csv2()`, `read_tsv()`.

`readxl`

* `excel_sheets()`: List all sheets in an Excel spreadsheet.
* `read_excel()`: Read xls and xlsx files.
    + Child functions: `read_xls()`, `read_xlsx()`

`utils`

* `data()`: Load specified data sets, or list the available data sets.
    + Use this function to load the data sets that accompany R packages, such as `openintro`'s `hsb2` and `email50` and `gapminder`'s `gapminder`.
* `read.table()`: Read a file in table format.
    + Child functions: `read.csv()`, `read.csv2()`, `read.delim()`, `read.delim2()`.

`XLConnect`

* `readWorksheetFromFile()`: Read data from worksheets in an Excel file.

***
## Import Data from the Internet

`httr`

* `GET()`: Get a URL.

`jsonlite`

* `read_json()`: Read and write JSON.

`readr`

* `read_delim()`: Read a delimited file (including csv & tsv) into a tibble.
    + Child functions: `read_csv()`, `read_csv2()`, `read_tsv()`.

`rjson`

* `fromJSON()`: Convert JSON to R.

`utils`

* `download.file()`: Download a file from the Internet.
    + Example of multiple files in the same link:
```{r, eval=FALSE}
url_rdata <- 
  "https://assets.datacamp.com/production/repositories/236/datasets/7f714f993f1ad4c3d26412ae1e537ce6355b1b54/iris.RData"

download.file(url_rdata, destfile = "datacamp.Rdata")

load("datacamp.Rdata")
```

* `unzip()`: Extract or list zip archives.
    + Example:
```{r, eval=FALSE}
download.file(
  "https://assets.datacamp.com/production/repositories/1069/datasets/578834f5908e3b2fa575429a287586d1eaeb2e54/countries2.zip",
  destfile = "Data Sets/Countries",
  mode = "wb")

unzip("Data Sets/Countries", exdir = "Data Sets")
```

***
## Import Data from a Database

`DBI`

* `dbBind()`: Bind values to a parameterized/prepared statement.
* `dbClearResult()`: Free all resources (local and remote) associated with a result set.
* `dbConnect()`: Connect to a DBMS.
* `dbDataType()`: Determine the SQL data type of an object.
* `dbDisconnect()`: Disconnect (close) a connection to a DBMS.
* `dbFetch()`: Fetch records from a previously executed query.
* `dbGetQuery()`: Send query, retrieve the results, and then clear result set.
* `dbListTables()`: List remote tables.
* `dbReadTable()`: Copy data frames to and from database tables.
* `dbSendQuery()`: Execute a query on a given database connection.
* `dbSendStatement()`: Execute a data manipulation statement on a given database connection.

*** 
## Reference Material

The `openintro` package contains data sets useful for practicing and teaching.