# Import

## Create Data

### Data Structures

#### Atomic Vector

`base`

* `c()`: Combine values into a vector or list.
* `factor()`: Create a factor variable.
* `seq()`: Generate a sequence.
* `vector()`: Create a vector.
    + This function is preferable to `c()` when creating an empty vector, according to the DataCamp article ["Five Tips to Improve Your R Code"](https://www.datacamp.com/community/tutorials/five-tips-r-code-improve).

#### Matrix

`base`

* `matrix()`: Create a matrix.

#### Array

`base`

* `array()`: Create an array.

#### List

`base`

* `c()`: Combine values into a vector or list.
* `list()`: Create a list.

`tibble`

* `tibble()`: Create a data frame or list.

#### Data Frame

`base`

* `data.frame`: Create a data frame.
    + Prefer `tibble::tibble` over `base::data.frame`.
* `list`: Create a list.

`tibble`

* `add_row`: Add rows to a data frame.
* `tibble`: Create a data frame or list.

#### References

* ["Vectors"](https://r4ds.had.co.nz/vectors.html) (Grolemund & Wickham, [*R for Data Science*](https://r4ds.had.co.nz/))
* ["Data structures"](http://adv-r.had.co.nz/Data-structures.html) (Wickham, [*Advanced R*](http://adv-r.had.co.nz/))

### Other

`stats`

* `rnorm()`: Generate a random normal distribution.

## Import Data from a Local Drive

`{base}`

* `attach`: Attach a set of R objects to the search path.
  * Allows objects in the database to be accessed by giving their names (e.g., `height` rather than `women$height`).
* `file.choose`: Choose a file interactively.
  * Use as `file = file.choose()` inside a data import function (such as `read_csv` and `readRDS`).
* `load`: Reload datasets saved with `save()`.
* `readRDS`: Restore an R object written with `saveRDS()`.

`{data.table}`

* `fread`: Similar to `read.table`, but faster and more convenient for large data sets.

`{foreign}`

* `read.spss`: Read an SPSS data file.

`{haven}`

* `read_sas`: Read and write SAS files.

`{readr}`

* `read_csv`: Read a comma separated value file into a tibble.
* `read_csv2`: Read a semi-colon separated value file into a tibble.
* `read_delim`: Read a delimited file into a tibble.
* `read_tsv`: Read a tab separated value file into a tibble.

`{readxl}`

* `excel_sheets`: List all sheets in an Excel spreadsheet.
* `read_excel`: Read xls and xlsx files.
* `read_xls`: Read a .xls file.
* `read_xlsx`: Read a .xlsx file.

`{utils}`

* `data`: Load specified data sets, or list the available data sets.
  * Use this function to load the data sets that accompany R packages, such as `openintro`'s `hsb2` and `email50` and `gapminder`'s `gapminder`.
* `read.csv`: Read a .csv file.
* `read.csv2`: Read semi-colon separated value file.
* `read.delim`: Read a delimited file.
* `read.delim2`: Read a delimited file where the decimal point is a comma.
* `read.table`: Read a file in table format.

`{XLConnect}`

* `readWorksheetFromFile`: Read data from worksheets in an Excel file.

## Import Data from the Internet

`{httr}`

* `GET`: Get a URL.

`{jsonlite}`

* `read_json`: Read and write JSON.

`{readr}`

* `read_csv`: Read a comma separated value file into a tibble.
* `read_csv2`: Read a semi-colon separated value file into a tibble.
* `read_delim`: Read a delimited file into a tibble.
* `read_tsv`: Read a tab separated value file into a tibble.

`{rjson}`

* `fromJSON`: Convert JSON to R.

`{utils}`

* `download.file`: Download a file from the Internet.
  * Example:
```{r, eval=FALSE}
download.file(
  "https://assets.datacamp.com/production/repositories/5028/datasets/a55843f83746968c7f118d82ed727db9c71e891f/snake_river_visits.rds",
  destfile = paste0(getwd(), "/Snake River Visits.rds"))

snake_river_visits <- readRDS(file.choose())
# or
path <- paste0(getwd(), "/Snake River Visits.rds")
snake_river_visits <- readRDS(path)
```

  * Example: Multiple files in the same link
```{r, eval=FALSE}
url_rdata <- 
  "https://assets.datacamp.com/production/repositories/236/datasets/7f714f993f1ad4c3d26412ae1e537ce6355b1b54/iris.RData"

download.file(url_rdata, destfile = "datacamp.Rdata")

load("datacamp.Rdata")
```

* `unzip`: Extract or list zip archives.
  * Example:
```{r, eval=FALSE}
download.file(
  "https://assets.datacamp.com/production/repositories/1069/datasets/578834f5908e3b2fa575429a287586d1eaeb2e54/countries2.zip",
  destfile = "Data Sets/Countries",
  mode = "wb")

unzip("Data Sets/Countries", exdir = "Data Sets")
```

## Import Data from a Database

`DBI`

* `dbBind()`: Bind values to a parameterized/prepared statement.
* `dbClearResult()`: Free all resources (local and remote) associated with a result set.
* `dbConnect()`: Connect to a DBMS.
* `dbDataType()`: Determine the SQL data type of an object.
* `dbDisconnect()`: Disconnect (close) a connection to a DBMS.
* `dbFetch()`: Fetch records from a previously executed query.
* `dbGetQuery()`: Send query, retrieve the results, and then clear result set.
* `dbListTables()`: List remote tables.
* `dbReadTable()`: Copy data frames to and from database tables.
* `dbSendQuery()`: Execute a query on a given database connection.
* `dbSendStatement()`: Execute a data manipulation statement on a given database connection.

## Reference Material

The `openintro` package contains data sets useful for practicing and teaching.