# Import or Create Data

## Create Data

`{base}`

* `array()`
* `c()`
  * See `base::vector()`.
* `data.frame()`
* `dir.create()`
* `factor()`
* `list()`
* `matrix()`
* `seq()`
* `vector()`
  * Preferable to `base::c()` when creating an empty vector (["Five Tips to Improve Your R Code"](https://www.datacamp.com/community/tutorials/five-tips-r-code-improve)).

`{stats}`

* `rnorm()`

`{tibble}`

* `add_row()`
* `tibble()`
* `tribble()`

## Local Drive

`{base}`

* `attach()`
  * Allows objects in the database to be accessed by giving their names (e.g., `height` rather than `women$height`).
* `file.choose()`
* `file.size()`
* `load()`
  * Reload datasets saved with `base::save()`.
* `readRDS()`
  * Restore an R object written with `base::saveRDS()`.
  
`{data.table}`

* `fread()`
  * Similar to `utils::read.table()`, but faster and more convenient for large data sets.
  
`{foreign}`

* `read.spss()`

`{haven}`

* `read_sas()`

`{readr}`

* `read_csv()`
* `read_csv2()`
* `read_delim()`
* `rread_tsv()`

`{readxl}`

* `excel_sheets()`
* `read_excel()`
* `read_xls()`
* `read_xlsx()`

`{utils}`

* `data()`
  * Load specified data sets, or list the available data sets.
  * Use this function to load the data sets that accompany R packages, such as `openintro::hsb2`, `openintro::email50`,and `gapminder::gapminder`.
* `read.csv()`
* `read.csv2()`
* `read.delim()`
* `read.delim2()`
* `read.table()`

`{XLConnect}`

* `readWorksheetFromFile()`

## Internet

`{httr}`

* `GET()`
  * Get a URL.

`{jsonlite}`

* `read_json()`

`{readr}`

* `read_csv()`
* `read_csv2()`
* `read_delim()`
* `read_tsv()`

`{rjson}`

* `fromJSON()`
  * Convert JSON to R.

`{utils}`

* `download.file()`
  * See example below.
* `unzip()`
  * See example below.

### Examples

`utils::download.file()`:

```{r, eval=FALSE}
download.file(
  "https://assets.datacamp.com/production/repositories/5028/datasets/a55843f83746968c7f118d82ed727db9c71e891f/snake_river_visits.rds",
  destfile = paste0(getwd(), "/Snake River Visits.rds")
)

# Option 1:
snake_river_visits <- readRDS(file.choose())

# Option 2:
snake_river_path   <- paste0(getwd(), "/Snake River Visits.rds")
snake_river_visits <- readRDS(snake_river_path)
```

`utils::download.file()` for .Rdata files:

```{r, eval=FALSE}
# Example 1:
download.file(
  "https://assets.datacamp.com/production/repositories/236/datasets/7f714f993f1ad4c3d26412ae1e537ce6355b1b54/iris.RData", 
  destfile = "datacamp_iris_dataset.Rdata"
)

load("datacamp_iris_dataset.Rdata")

# Example 2:
download.file(
  "https://assets.datacamp.com/production/repositories/235/datasets/3b6fc2923b599058584b57d8c605c6bef454d273/CHIS2009_reduced_2.Rdata",
  destfile = "chis_2009.Rdata",
  # The documentation for `download.file` indicates that the function will
  # automatically include `mode = "wb"` for .Rdata files. That may have happened
  # in Example 1, but didn't happen in Example 2, which is why I've included it.
  mode = "wb"
)

load("chis_2009.Rdata")
```

`utils::unzip()`:

```{r, eval=FALSE}
download.file(
  "https://assets.datacamp.com/production/repositories/1069/datasets/578834f5908e3b2fa575429a287586d1eaeb2e54/countries2.zip",
  destfile = "Data Sets/Countries",
  mode = "wb"
)

unzip("Data Sets/Countries", exdir = "Data Sets")
```

## Database

`{DBI}`

* `dbBind()`
* `dbClearResult()`
* `dbConnect()`
* `dbDataType()`
* `dbDisconnect()`
* `dbExecute()`
* `dbFetch()`
* `dbGetQuery()`
* `dbListTables()`
* `dbReadTable()`
* `dbSendQuery()`
* `dbSendStatement()`
* `Id()`
  * See example below.
* `SQL()`
  * See example below.

`{dbplyr}`

* `in_schema()`
* `tbl()`

`{odbc}`

* `dbConnect()`
  * See example below for specifying a database.
  * See example below for troubleshooting hang-ups.

### Examples

`DBI::Id()` vs `DBI::SQL()`:

The following are comparable methods of accessing `<schema_name>.<table_name>`:

```{r, eval=FALSE}
dbCreateTable(
  conn,
  Id(schema = "schema_name", table = "table_name"),
  fields = "field_name"
)

dbCreateTable(
  conn,
  name   = SQL("schema_name.table_name"),
  fields = "field_name"
)
```

`odbc::dbConnect()`:

```{r, eval=FALSE}
# Specify database:
conn_1 <- dbConnect(odbc::odbc(), dsn = "sql") # Default database: Actuary

conn_2 <- dbConnect(odbc::odbc(), dsn = "sql", Database = "Staging")

# Troubleshoot hang-ups
# It appears that RStudio's attempt to load information into the Connection Pane,
# via `odbc::dbConnect`, can sometimes cause the call to hang, indefinitely.
# Use the following code to access the database in such a situation:

options(connectionObserver = FALSE)
```

### References

* ["Databases using R"](db.rstudio.com)

### SQL and R

#### Tips and Troubleshooting Hints

* Temp Tables
  * Use `SET NOCOUNT ON` in the SQL query to prevent R confusing the SQL rowcount return value with the temporary table.
* Try case-sensitive syntax (e.g., when querying a database other than the default).
* Try selecting specific columns, rather than `SELECT *`, as SQL BLOB columns cause trouble. 

#### SQL References

* ["10 Easy Steps to a Complete Understanding of SQL"](https://blog.jooq.org/2016/03/17/10-easy-steps-to-a-complete-understanding-of-sql/)
* ["Query Planning"](https://www.sqlite.org/queryplanner.html)

## Notes & References

* ["Data structures"](http://adv-r.had.co.nz/Data-structures.html) (Wickham, [*Advanced R*](http://adv-r.had.co.nz/))
* `{openintro}` contains data sets useful for practicing and teaching.
* ["Vectors"](https://r4ds.had.co.nz/vectors.html) (Grolemund & Wickham, [*R for Data Science*](https://r4ds.had.co.nz/))