# Program

***
"Surrounding [the tools for importing, tidying, transforming, visualising, modeling, and communicating data] is programming. Programming is a cross-cutting tool that you use in every part of a project. You don't need to be an expert programmer to be a data scientist, but learning more about programming pays off because becoming a better programmer allows you to automate common tasks, and solve new problems with greater ease."  
- Garrett Grolemund & Hadley Wickham, *R for Data Science*

***
## Conditionals & Control Flows

`base`

* Control (access documentation using "?Control")
    + `if (cond) expr`
        + The key difference between `if (cond) expr` and `ifelse` is that `if (cond) expr` will evaluate only the first element of an object with `length > 1`. See the documentation for each function and the ateucher's GitHub example titled ["R: if vs ifelse"](https://gist.github.com/ateucher/c7359f566eded9fcd4a255f4cbd4fe67) to learn more.
    + `for (var in seq) expr`
    + `while (cond) expr`
    + `repeat expr`
    + `break`
    + `next`
* `identical()`: Test objects for exact equality.
    + Use `identical()` rather than `==` and `!=` in `if` and `while` statements to test for equality.
* 'ifelse()`: Conditional element selection.
    + `dplyr::if_else()` is more strict by checking the object type.
* `stop()`: Stop execution of the expression and execute an error action.
    + Useful in combination with an `if` statement when you want to generate helpful error messages.
* `stopifnot()`: Ensure the truth of an R expression.
    + Prefer `base::stop()` to `base::stopifnot()`.

`dplyr`

* `case_when()`: A general vectorized `if`.
* `if_else()`: Vectorized `if`.

***
## Environment and Workspace

`{base}`

* `dir`: List the files in a directory/folder.
* Environments
    + `baseenv`: The environment of the `base` package, it's enclosing environment ("parent environment") is the empty environment.
    + `emptyenv`: The empty environment, which is the ancestor of all environments and the only environment without an enclosing environment.
    + `environment`: The current environment.
    + `globalenv`: The environment in which you normally work, it's enclosing environment is the last package attached with `library` or `require`.
    + `new.env`: Create a new environment.
* `exists`: Look for an R object of the given name and possibly return it.
    + Must use quotations to name the object.
* `getOption`: Set and examine global options.
* `getwd`: Get the working directory.
* `history`: Display the previous 25 commands.
* `install.packages`: Install packages from repositories or local files.
* `library`: Load and attach packages, returning an error if the packages does not exist.
* `list.files`: List the files in a directory/folder.
* `loadedNamespaces: Return the loaded name spaces.
* `loadhistory`: Recall command history.
* `ls`: List objects in the specified environment.
* `options`: Set and examine global options.
* `q`: Terminate an R session.
* `R.version`: Version information.
* `R.version.string`: Version information. 
  * Same call as `R.version$version.string`.
* `require`: Load and attach packages, returning `FALSE` if the package does not exist.
* `rm`: Remove objects from a specified environment.
* `savehistory`: Save command history (default value is ".Rhistory").
* `save.image`: Save the current workspace.
* `search`: Return a list of attached packages and R objects.
* `searchpaths`: Return the path to attached packages.
* `setwd()`: Set the working directory file path.
  * When using Windows, use "/" instead of "\".
* `Sys.info`: Extract system and user information.
  * Example: `Sys.info()[c("sysname", "release")]`.
    
`{gdata}`

* `object.size`: Report the space allocated for an object.
  * See also `utils::object.size`.

`{installr}`

* `updateR`: Check for the latest R version; downloads and installs new R versions.

`{pryr}`

* `where`: Find where a name is defined.

`{utils}`

* `object.size`: Report the space allocated for an object.
  * See also `gdata::object.size`.
* `sessionInfo`: Collect information about the current R session.
    
References:

* ["Environments"](http://adv-r.had.co.nz/Environments.html#environments) (Hadley Wickham, [*Advanced R*](http://adv-r.had.co.nz/))

***
## Evaluation (Standard and Non-standard)

`base`

* `cat()`: Concatenate and print.
* `print(): Print the argument to the Console.
    + A shortcut to `print()` is to place the code you want printed inside parentheses.
* `quote()`: Return the argument, unevaluated.
* `writeLines()`: Display quotes and backslashes as they would be read, rather than as R stores them (i.e., see the raw contents of the string, as the `print()` representation is not the same as the string itself).

`rlang`

* Quosures
    + `enquo()`, `new_quosure()`, `quo()`.
    
References:

* ["Non-standard evaluation"](http://adv-r.had.co.nz/Computing-on-the-language.html) (Hadley Wickham, [*Advanced R*](http://adv-r.had.co.nz/))
* ["Non-standard evaluation"](https://cran.r-project.org/web/packages/lazyeval/vignettes/lazyeval.html) (Hadley Wickham, `lazyeval` package vignette)
* ["Programming with dplyr"](https://dplyr.tidyverse.org/articles/programming.html) (dplyr.tidyverse.org)

***
## Functionals

`{base}`

* Apply Functions
  * `apply`: Apply functions over array margins.
  * `lapply`: Apply a function over a list or vector.
  * `sapply`: Apply a function over a list or vector and return a vector or matrix.
  * `vapply`: A safer version of `sapply`, as it requires the output type to be predetermined.
  * `mapply`: Apply a function to multiple list or vector arguments.
  * `rapply`: Recursively apply a function to a list. 
  * `tapply`: Apply a function over a ragged array.

`{purrr}`

* `map`: Apply a function to each element of a vector.
  * `tidyr::unnest` is useful in changing the list-column output of `map` into rows.
* `map2`: Map over multiple inputs simultaneously.
* `map_if`: Apply a function to elements of that match a condition.
* `possibly`: Usese a default value whenever an error occurs.
* `quietly`: Capture side effects in a list with components `result`, `output`, `messages`, and `warnings`.
* `safely`: Capture side effects in a list with components `result` and `error`.
* `transpose`: Transpose a list (turn a list-of-lists inside-out).

***
## Functions

`{assertive}`

* `assert_*`: Check whether the input is `*` (e..g, `assert_is_numeric`) and throw and error if the input does not meet the condition.
* `coerce_to`: Coerce the input to a different class, with a warning.
* `is_*`: Checks whether the input matches the condition specified by `*` (e.g., `is_non_positive`).
* `use_first`: Use only the first element of a vector.

`{base}`

* `do.call`: Execute a function call from a name or a function and a list of arguments to be passed to the function.
* `invisible`: Return a (temporarily) invisible copy of an object.
* `match.arg`: Argument verification.
  * Useful when matching a character argument specified in the function signature. For example, 

```{r}
args(prop.test)

# The body of `prop.test` contains the following line of code:
# `alternative <- match.arg(alternative), which reassigns it to the selected
# character vector.
```
  
* `message`: Generage a diagnostic message.
  * Preferable to generating a message using `cat`.
* `return`: Return a value from a function.
  * Useful in `if` statements where one condition is simple and the other is complex (see section 19.6.1 "Explicit return statements" in Hadley Wickham's [*R for Data Science*](https://r4ds.had.co.nz/functions.html).
* `setNames`: Set the names in an object.
  * Useful in function writing; see documentation.
* `stop`: Stop execution of the expression and execute an error action.
  * Useful in combination with an `if` statement when you want to generate helpful error messages.
* `stopifnot`: Ensure the truth of an R expression.
  * See section 19.5.2 "Checking values" in Hadley Wickham's [*R for Data Science*](https://r4ds.had.co.nz/functions.html) for a discussion of `stop` versus `stopifnot`.
  * Consider functions from `{assertive}` as an alternative to `stopifnot` and `stop`.
* `unlist`: Flatten lists.
  * Useful when using `purrr`'s `map` functions, which return objects as type `list`.

`{zeallot}`

* `%<-%`: Multiple assignment operator.
  * Useful when you want to return multiple output from a function. For example:
  
```{r, eval=FALSE}
session <- function() {
  
  list(
    r_version = R.version.string,
    operating_system = Sys.info()[c("sysname", "release")],
    loaded_pkgs = loadedNamespaces()
  )
}

c(vrsn, os, pkgs) %<-% session() 
```

***
## Learn About an Object

`base`

* `args()`: Display the argument names and default values of a function.
* `attributes()`: View or assign an objects attributes (e.g., `class()`, `dim()`, `dimnames()`, `names()`, `row.names()`). 
* `body()`: Get or set the body of a function.
* `colnames()`: Retrieve or set column names.
* `dim()`: Retrieve or set the dimnames of an object.
* `dimnames()`: Retrieve or set the dimension names of an object.
* `formals()`: Get or set the formal arguments of a function.
* `help()`: Get the topic documentation.
* `help.search()`: Search the help system for documentation matching a given character string.
* `vignette()`: View a specified package vignette.
* ?object_name
* ??object_name
* `rownames()`: Retrieve or set row names.

***
## Loops

`base`

* `seq()`: Sequence generation.
    + This functions makes `length()` unnecessary.
    + Child function: `seq_along()`
        + In `for` loops, safer than using `ncol()` or `nrow()`.

***
## Optimization

`microbenchmark`

* `microbenchmark()`: Sub-millisecond accurate timing of expression evaluations.
    + A more accurate replacement of `system.time(replicate(1000, expr))`.

***
## Pipes

`magrittr`: Forward-pipe operator for R.

* `freduce()`: Apply a list of functions sequentially.
* `%<>%`: Compound assignment-pipe operator.
* `%>%`: Forward-pipe operator.
* `%$%`: Expositions-pipe operator.

***
## Popups

`svDialogs`

* `dlg_message`: Display a modal message box (works in Windows, MacOS, and Linux).

`tcltk`

* `tk_messageBox`: Display a generic message box using Tk (Windows-specific).

## Selecting & Subsetting

`base`

* `subset()`: Subset vectors, matrices, and data frames.

`dplyr`

* `first()`: Extract the first element of a vector.
* `last()`: Extract the last element of a vector.
* `nth(): Extract the nth element of a vector.
* `select()`: Select/rename variables. 
    + Helper functions include: `starts_with()`, `ends_with()`, `contains()`, `matches()`, `num_range()`, and `one_of()`.
    + A closely-related function is `dplyr::rename()`.
* `.$variable_name`: Select a variable from a data frame in a pipe.
    + Example:
```{r, eval=FALSE}
ui_summary_table <- 
  aws_vendors %>% 
  filter(str_detect(vendor_name, "UTAH INTERACTIVE")) %>% 
  .$vendor_id %>% 
  map(query_summary_table) %>% 
  bind_rows()
```
* `.[["variable_name"]]`: Select a vairable from a data frame in a pipe.
    + Example:
```{r, eval=FALSE}
  odbc_aws %>% 
    dbGetQuery(
      paste("
            SELECT id
            FROM batch
            WHERE entity_id = ", t_id, "
            AND status IN ('PROCESSED', 'PROCESSING')")) %>% 
    .[["id"]] %>% 
  as.double()
```

References:

* ["Indexing lists in #rstats. Inspired by Residence Inn"](https://twitter.com/hadleywickham/status/643381054758363136) (Hadley Wickham, Twitter, 14 September 2015)

***
## Version Control

Git

* [Git](https://git-scm.com/)
* [*Pro Git*](https://git-scm.com/book/en/v2) by Scott Chacon and Ben Straub
* [*Git and GitHub*](http://r-pkgs.had.co.nz/git.html#git-learning) by Hadley Wickham
* [*Happy Git and GitHub for the useR*](http://happygitwithr.com/) by Jenny Bryan

* `git branch`: List, create, or delete branches.
    + `git branch -d <branch_name>`: Delete a local branch.
        + See ["Delete branch in RStudio pop-up"](https://community.rstudio.com/t/delete-branch-in-rstudio-pop-up/15465) for help removing branches in RStudio after removing them from Git.

`{packrat}`

* See note on the `packrat` package in the "Referenced Packages" section.
* `snapshot`: Capture and store the packages and versions in use.
* `restore`: Load the most recent snapshot to the project's private library.