# Program

"Surrounding [the tools for importing, tidying, transforming, visualising, modeling, and communicating data] is programming. Programming is a cross-cutting tool that you use in every part of a project. You don't need to be an expert programmer to be a data scientist, but learning more about programming pays off because becoming a better programmer allows you to automate common tasks, and solve new problems with greater ease."  
- Garrett Grolemund & Hadley Wickham, *R for Data Science*

## Characters

References:

- `?Quotes`

## Conditionals & Control Flows

`{base}`

- Control (access documentation using `?Control`)
  - `if (cond) expr`
    - The key difference between `if (cond) expr` and `ifelse` is that `if (cond) expr` will evaluate only the first element of an object with `length > 1`. See the documentation for each function and the ateucher's GitHub example titled ["R: if vs ifelse"](https://gist.github.com/ateucher/c7359f566eded9fcd4a255f4cbd4fe67) to learn more.
  - `for (var in seq) expr`
  - `while (cond) expr`
  - `repeat expr`
  - `break`
  - `next`
- `identical()`: Test objects for exact equality.
  - Use `identical()` rather than `==` and `!=` in `if` and `while` statements to test for equality.
- `ifelse()`: Conditional element selection.
  - `dplyr::if_else()` is more strict by checking the object type.
- `stop()`: Stop execution of the expression and execute an error action.
  - Useful in combination with an `if` statement when you want to generate helpful error messages.
- `stopifnot()`: Ensure the truth of an R expression.
  - Prefer `base::stop()` to `base::stopifnot()`.

`dplyr`

- `case_when()`: A general vectorized `if`.
- `if_else()`: Vectorized `if`.

## Environment and Workspace

`{base}`

- `.libPaths()`
- `baseenv()`
  - The environment of the `{base}` package; it's enclosing environment ("parent environment") is the empty environment.
- `dir()`
  - List the files in a directory.
- `dir.create()`
  - Create a file path in the computer's file system.
- `emptyenv()`
  - The empty environment, which is the ancestor of all environments and the only environment without an enclosing environment.
- `environment()`
  - The current environment.
- `environmentName()`
- `exists()`
  - Remember that `R` will look for an object in parent environments until it reaches the empty environment, so use `inherits = FALSE` to limit the search to only the current environment.
- `getOption()`
  - Set and examine global options.
- `getwd()`
  - Get the working directory.
- `globalenv()`
  - The environment in which you normally work, it's enclosing environment is the last package attached with `library()` or `require()`.
- `history()`
  - Display the previous 25 commands.
- `install.packages()`
  - Install packages from repositories or local files.
- `library()`
  - Load and attach packages, returning an error if the packages does not exist.
- `list.files()`
  - List the files in a directory/folder.
- `loadedNamespaces()`
  - Return the loaded name spaces.
- `loadhistory()`
  - Recall command history.
- `ls()`
  - List objects in the specified environment.
- `list2env()`
  - From a list, build or add to an environment.
- `new.env()`
  - Create a new environment.
- `options()`
  - Set and examine global options.
- `parent.env()`
  - Return the enclosing environment of the environment listed as an argument.
  - `base::parent.env()` returns information that can be unhelpful, so use with `base::environmentName()`, as follows: `parent.env(environment_name) %>% environmentName()`.
- `q()`
  - Terminate an R session.
- `R.version()`
  - Version information.
- `R.version.string()`
  - Version information. 
  - Same call as `R.version$version.string()`.
- `require()`
  - Load and attach packages, returning `FALSE` if the package does not exist.
- `rm()`
  - Remove objects from a specified environment.
- `savehistory()`
  - Save command history (default value is ".Rhistory").
- `save.image()`
  - Save the current workspace.
- `search()`
  - Return a list of attached packages and R objects.
- `searchpaths()`
  - Return the path to attached packages.
- `setwd()`
  - Set the working directory file path.
  - When using Windows, use "/" instead of "\".
- `Sys.getenv()`
  - See also `Sys.setenv()`.
- `Sys.info()`
  - Extract system and user information.
  - Example: `Sys.info()[c("sysname", "release")]`.
- `Sys.setenv()`
  - See also `Sys.setenv()`.

`{gdata}`

- `object.size()`
  - Report the space allocated for an object.
  - See also `utils::object.size()`.

`{here}`

- `here()`

`{installr}`

- `updateR()`
  - Check for the latest R version; downloads and installs new R versions.

`{pryr}`

- `where()`: Find where a name is defined.


`{utils}`

- `ls.str()`: List objects and  their structure.
- `object.size()`: Report the space allocated for an object.
  - See also `gdata::object.size()`.
- `sessionInfo()`: Collect information about the current R session.
    
References:

- ["Environments"](http://adv-r.had.co.nz/Environments.html#environments) (Hadley Wickham, [*Advanced R*](http://adv-r.had.co.nz/))

## Evaluation (Standard and Non-standard)

`base`

- `cat()`
  - Concatenate and print.
- `print()
  - Print the argument to the Console.
  -  A shortcut to `print()` is to place the code you want printed inside parentheses.
- `quote()`
  - Return the argument, unevaluated.
- `writeLines()`
  - Display quotes and backslashes as they would be read, rather than as R stores them (i.e., see the raw contents of the string, as the `print()` representation is not the same as the string itself).

`rlang`

- Quosures
  - `enquo()`, `new_quosure()`, `quo()`.
    
References:

- ["Non-standard evaluation"](http://adv-r.had.co.nz/Computing-on-the-language.html) (Hadley Wickham, [*Advanced R*](http://adv-r.had.co.nz/))
- ["Non-standard evaluation"](https://cran.r-project.org/web/packages/lazyeval/vignettes/lazyeval.html) (Hadley Wickham, `lazyeval` package vignette)
- ["Programming with dplyr"](https://dplyr.tidyverse.org/articles/programming.html) (dplyr.tidyverse.org)

## Functionals

`{base}`

- Apply Functions
  - `apply`
    - Apply functions over array margins.
  - `lapply`
    - Apply a function over a list or vector.
  - `sapply`
    - Apply a function over a list or vector and return a vector or matrix.
  - `vapply`
    - A safer version of `sapply`, as it requires the output type to be predetermined.
  - `mapply`
    - Apply a function to multiple list or vector arguments.
  - `rapply`
    - Recursively apply a function to a list. 
  - `tapply`
    - Apply a function over a ragged array.

`{purrr}`

- `map`
  - Apply a function to each element of a vector.
  - `tidyr::unnest` is useful in changing the list-column output of `map` into rows.
- `map2`
  - Map over multiple inputs simultaneously.
- `map_if`
  - Apply a function to elements of that match a condition.
- `possibly`
  - Uses a default value whenever an error occurs.
- `quietly`
  - Capture side effects in a list with components `result`, `output`, `messages`, and `warnings`.
- `safely`
  - Capture side effects in a list with components `result` and `error`.
- `transpose`
  - Transpose a list (turn a list-of-lists inside-out).

## Functions

`{assertive}`

- `assert_*()`
  - Check whether the input is `*` (e..g, `assert_is_numeric()`) and throw and error if the input does not meet the condition.
- `coerce_to()`
  - Coerce the input to a different class, with a warning.
- `is_*()`
  - Checks whether the input matches the condition specified by `*` (e.g., `assertive::is_non_positive()`).
- `use_first()`
  - Use only the first element of a vector.

`{base}`

- `do.call()`
  - Execute a function call from a name or a function and a list of arguments to be passed to the function.
- `invisible()`
  - Return a (temporarily) invisible copy of an object.
- `match.arg()`
  - Argument verification. Useful when matching a character argument specified in the function signature. See example.
- `message()`
  - Generate a diagnostic message.
  - Preferable to generating a message using `cat`.
- `return()`
  - Return a value from a function.
  - Useful in `if` statements where one condition is simple and the other is complex (see section 19.6.1 "Explicit return statements" in Hadley Wickham's [*R for Data Science*](https://r4ds.had.co.nz/functions.html).
- `setNames()`
  - Set the names in an object.
  - Useful in function writing; see documentation.
- `stop()`
  - Stop execution of the expression and execute an error action.
  - Useful in combination with an `if` statement when you want to generate helpful error messages.
- `stopifnot()`
  - Ensure the truth of an R expression.
  - See section 19.5.2 "Checking values" in Hadley Wickham's [*R for Data Science*](https://r4ds.had.co.nz/functions.html) for a discussion of `stop()` versus `stopifnot()`.
  - Consider functions from `{assertive}` as an alternative to `stopifnot` and `stop`.
- `unlist()`
  - Flatten lists.
  - Useful when using `purrr`'s `map` functions, which return objects as type `list`.
  
`{zeallot}`

- `%<-%`
  - Multiple assignment operator (see example below).
  
### Examples

`base::args()`

```{r}
args(prop.test)

# The body of `prop.test` contains the following line of code:
# `alternative <- match.arg(alternative), which reassigns it to the selected
# character vector.
```

`zeallot::%<-%`:

```{r, eval=FALSE}
session <- function() {
  
  list(
    r_version        = R.version.string,
    operating_system = Sys.info()[c("sysname", "release")],
    loaded_pkgs      = loadedNamespaces()
  )
}

c(vrsn, os, pkgs) %<-% session() 
```

## Learn About an Object

- `?object_name`
- `??object_name`

`{base}`

- `args()`
- `attributes()`
  - View or assign an objects attributes (e.g., `class()`, `dim()`, `dimnames()`, `names()`, `row.names()`). 
- `body()`
  - Get or set the body of a function.
- `colnames()`
- `dim()`
  - Retrieve or set the dimnames of an object.
- `dimnames()`
  - Retrieve or set the dimension names of an object.
- `formals()`
  - Get or set the formal arguments of a function.
- `help()`
  - Get the topic documentation.
- `help.search()`
  - Search the help system for documentation matching a given character string.
- `vignette()`
- `rownames()`

## Loops

- `base::seq()`: Sequence generation (this functions makes `length()` unnecessary).
- `base::seq_along()`: In `for` loops, safer than using `base::ncol()` or `base::nrow()`.

## Optimization

`microbenchmark::microbenchmark()`: Sub-millisecond accurate timing of expression evaluations.
  - A more accurate replacement of `system.time(replicate(1000, expr))`.

## Popups

- `svDialogs::dlg_message()`: Display a modal message box (works in Windows, MacOS, and Linux).
- `tcltk::tk_messageBox()`: Display a generic message box using Tk (Windows-specific).

## Selecting, Subsetting, and Piping

[`{base}`](https://rdocumentation.org/packages/base/versions/3.6.2)

- `.$variable_name`: See example below.
- `.[["variable_name"]]`: See example below.

[`{dplyr}`](https://rdocumentation.org/packages/dplyr/versions/0.7.8)

- `pull()`
  - See `magrittr::extract()`, `purrr::pluck()`.

[`{magrittr}`](https://rdocumentation.org/packages/magrittr/versions/2.0.3)

- `extract()`
  - See `dplyr::pull()`, `purrr::pluck()`.

### Examples

- `base::.$variable_name`:

```{r, eval=FALSE}
ui_summary_table <- 
  aws_vendors %>% 
  filter(str_detect(vendor_name, "UTAH INTERACTIVE")) %>% 
  .$vendor_id %>% 
  map(query_summary_table) %>% 
  bind_rows()
```

- `base::.[["variable_name"]]`:

```{r, eval=FALSE}
odbc_aws %>% 
  dbGetQuery(
    paste("
          SELECT id
          FROM batch
          WHERE entity_id = ", t_id, "
          AND status IN ('PROCESSED', 'PROCESSING')"
    )
  ) %>% 
  .[["id"]] %>% 
  as.double()
```

### References

- ["Indexing lists in #rstats. Inspired by Residence Inn"](https://twitter.com/hadleywickham/status/643381054758363136) (Hadley Wickham, Twitter, 14 September 2015)

## Style Guide

- ["The Tidyverse Style Guide"](https://style.tidyverse.org/) by Hadley Wickham.
- General Layout and Ordering (taken from the outdated ["Google's R Style Guide"](http://web.stanford.edu/class/cs109l/unrestricted/resources/google-style.html))
  - Title
  - Author
  - File description (e.g., purpose of program, inputs, outputs)
  - `source()` and `library()` statements
  - Function definitions
  - Executed statements.

## System Commands

`{base}`

- `shell()`
- `system()`
- `system2()`

`{sys}`

- `exec()`
- `exec_r()`

## Version Control

- `base::update.packages()`
  - 00LOCK error when updating a package: Use `update.packages(ask = FALSE, checkBuilt = TRUE, INSTALL_opts = "--no-lock")`
- `installr::updateR()`
  - Remove old versions from Windows:
    - Control Panel > All Control Panel Items > Programs and Features > uninstall old versions
    - C: > Program Files > R > delete old versions.
- `packrat::snapshot()`: Capture and store the packages and versions in use.
- `packrat::restore()`: Load the most recent snapshot to the project's private library.
- Packrat still seems to be under development. The idea sounds good, but in practice the packrat package has caused lots of problems. 

#### Git

- [*A successful Git branching model*](https://nvie.com/posts/a-successful-git-branching-model/)
- [git-scm.com](https://git-scm.com/)
- [*Git and GitHub*](http://r-pkgs.had.co.nz/git.html#git-learning) by Hadley Wickham
- [*Happy Git and GitHub for the useR*](http://happygitwithr.com/) by Jenny Bryan
- [*Pro Git*](https://git-scm.com/book/en/v2) by Scott Chacon and Ben Straub
- [*Understanding the GitHub flow*](https://guides.github.com/introduction/flow/)
- `git branch`: List, create, or delete branches.
  - `git branch -d <branch_name>`: Delete a local branch.
    - See ["Delete branch in RStudio pop-up"](https://community.rstudio.com/t/delete-branch-in-rstudio-pop-up/15465) for help removing branches in RStudio after removing them from Git.
- Git and R Projects
  - R Projects (.Rproj files) should not be nested. Doing so causes problems with Git when tracking changes to a child .Rproj file.