--- 
title: "R Reference Manual"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
always_allow_html: yes
---
# Introduction

The purpose of this manual is to increase my ability to use R well, which means that I must be able to get things done. My vision is that this manual will allow me to document the material I learn in a way that will help me work more effectively and efficiently. Most of the documentation here will link to existing online material so that my notes supplement rather than duplicate.

In order to help me reference material in the context I would use it, I organized the material according to Garrett Grolemund and Hadley Wickham's conception of the tools needed to tackle about 80% of the tasks required in a typical data science project ("[Introduction](https://r4ds.had.co.nz/introduction.html)", *R for Data Science*). Those tools are: import, tidy, transform, visualize, model, communicate, and program.

```{r, echo=FALSE}
library(DiagrammeR)

grViz(" digraph {

graph [rankdir = LR;
       compound = TRUE;
       nodesep = 1]

subgraph cluster0 {
  style = rounded;
  label = Program;
  labeljust = center;
  node [color = white];

  Import; Tidy; Communicate

  subgraph cluster1 {
    style = rounded;
    bgcolor = lightgrey;
    label = Understand;
    labeljust = center;
    node [color = lightgrey];

    Transform -> Visualize -> Model -> Transform;

    {rank = same; Model; Visualize}
  }
  
  Import -> Tidy -> Transform;

  edge [headport = w]
  {Model; Visualize} -> Communicate [ltail = cluster1]
}
}")
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```