--- 
title: "R Reference Manual"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
always_allow_html: yes
---

# Introduction

In *R for Data Science*, Garrett Grolemund and Hadley Wickham outline the tools needed to tackle about 80% of the tasks required in a typical data science project ("[Introduction](https://r4ds.had.co.nz/introduction.html)"). Those tools,  illustrated in the diagram below and which are the organizing template for this reference manual, are: 

- Import
- Tidy
- Transform
- Visualize
- Model
- Communicate
- Program

```{r, echo=FALSE}
DiagrammeR::grViz(" 

digraph {
  graph [rankdir = LR; compound = TRUE; nodesep = 1]
  
  subgraph cluster0 {
    style = rounded;
    label = Program;
    labeljust = center;
    node [color = white];
  
    Import; Tidy; Communicate
  
    subgraph cluster1 {
      style = rounded;
      bgcolor = lightgrey;
      label = Understand;
      labeljust = center;
      node [color = lightgrey];
  
      Transform -> Visualize -> Model -> Transform;
  
      {rank = same; Model; Visualize};
      
    }
    
    
    Import -> Tidy -> Transform
    
  
    edge [headport = w]
    {Model; Visualize} -> Communicate [ltail = cluster1]
  }
}")
```

## References & Resources

- For an introduction to the R programming language, see the R Project for Statistical Computing's ["What is R?"](https://www.r-project.org/about.html) and Wikipedia's ["R (programming language)."](https://en.wikipedia.org/wiki/R_(programming_language))

- To download R, go to [r-project.org](https://www.r-project.org/) and choose the cloud CRAN Mirror option.

- To program in the R language on a user-friendly platform, download the [RStudio](https://www.rstudio.com/) IDE.

- [The R Project for Statistical Computing](https://www.r-project.org/)
  - [Library of R Packages](https://cran.r-project.org/web/packages/)
  - [*Getting Help with R*](https://www.r-project.org/help.html)
  - [*The R Manuals*](https://cran.r-project.org/manuals.html)
  - [*Frequently Asked Questions*](https://cran.r-project.org/faqs.html)
  - [*Books Related to R*](https://www.r-project.org/doc/bib/R-books.html)
  - [*Documentation*](https://www.r-project.org/other-docs.html)

* [RStudio](https://www.rstudio.com/)
  - [*RStudio Cheat Sheets*](https://www.rstudio.com/resources/cheatsheets/)
  - [*Webinars and Videos On Demand*](https://www.rstudio.com/resources/webinars/)
  - [*Online learning*](https://www.rstudio.com/online-learning/)
  - [*RStudio Blog*](https://blog.rstudio.com/)

* Online Manuals
  - [*R for Data Science*](http://r4ds.had.co.nz/)
  - [*Advanced R*](https://adv-r.hadley.nz) by Hadley Wickham
  - [*R Packages*](http://r-pkgs.had.co.nz/) by Hadley Wickham
  - [*The R Inferno*](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf) by Patrick Burns
  - [*The tidyverse style guide*](http://style.tidyverse.org/)
  - [*Efficient R Programming*](https://bookdown.org/csgillespie/efficientR/)
  - [*Free R Reading Material*](https://committedtotape.shinyapps.io/freeR/)
  - [*What They Forgot to Teach You About R*](https://rstats.wtf/) by Jenny Bryan and Jim Hester

- Other Online Resources
  - [DataCamp](https://www.datacamp.com/)
  - [RDocumentation](https://www.rdocumentation.org/)
  - [R Bloggers](https://www.r-bloggers.com/how-to-learn-r-2/)
    - ["Tutorials for learning R"](https://www.r-bloggers.com/how-to-learn-r-2/)
  - [Regular Expressions 101](https://regex101.com/)

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
