--- 
title: "R Reference Manual"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
always_allow_html: yes
---
# Introduction

In *R for Data Science*, Garrett Grolemund and Hadley Wickham outline the tools needed to tackle about 80% of the tasks required in a typical data science project ("[Introduction](https://r4ds.had.co.nz/introduction.html)"). Those tools, illustrated in the diagram below and the organizing template for this reference manual, are: 

* Import
* Tidy
* Transform
* Visualize
* Model
* Communicate
* Program

```{r, echo=FALSE}
library(DiagrammeR)

grViz(" digraph {

graph [rankdir = LR;
       compound = TRUE;
       nodesep = 1]

subgraph cluster0 {
  style = rounded;
  label = Program;
  labeljust = center;
  node [color = white];

  Import; Tidy; Communicate

  subgraph cluster1 {
    style = rounded;
    bgcolor = lightgrey;
    label = Understand;
    labeljust = center;
    node [color = lightgrey];

    Transform -> Visualize -> Model -> Transform;

    {rank = same; Model; Visualize}
  }
  
  Import -> Tidy -> Transform;

  edge [headport = w]
  {Model; Visualize} -> Communicate [ltail = cluster1]
}
}")
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

<!--chapter:end:index.Rmd-->

# Import
***

## Create Data
***

* `base::array`
* `base::c`
* `base::data.frame`
* `base::dir.create`
  * Create a file path in the computer's file system.
* `base::factor`
* `base::list`
* `base::matrix`
* `base::seq`
* `base::vector`
  * Preferable to `base::c` when creating an empty vector (["Five Tips to Improve Your R Code"](https://www.datacamp.com/community/tutorials/five-tips-r-code-improve)).
* `stats::rnorm`
* `tibble::add_row`
* `tibble::tibble`
* `tibble::tribble`
  * Row-wise tibble creation.

## Import Data from a Local Drive
***

* `base::attach`
  * Attach a set of R objects to the search path.
  * Allows objects in the database to be accessed by giving their names (e.g., `height` rather than `women$height`).
* `base::file.choose`
  * Use as `file = file.choose()` inside a data import function (such as `read_csv` and `readRDS`).
* `base::load`
  * Reload datasets saved with `base::save`.
* `base::readRDS`
  * Restore an R object written with `base::saveRDS`.
* `data.table::fread`
  * Similar to `utils::read.table`, but faster and more convenient for large data sets.
* `foreign::read.spss`
* `haven::read_sas`
* `readr::read_csv`
* `readr::read_csv2`
* `readr::read_delim`
* `readr::read_tsv`
* `readxl::excel_sheets`
* `readxl::read_excel`
* `readxl::read_xls`
* `readxl::read_xlsx`
* `utils::data`
  * Load specified data sets, or list the available data sets.
  * Use this function to load the data sets that accompany R packages, such as `openintro::hsb2`, `openintro::email50`,and `gapminder::gapminder`.
* `utils::read.csv`
* `utils::read.csv2`
* `utils::read.delim`
* `utils::read.delim2`
* `utils::read.table`
  * Read a file in table format.
* `XLConnect::readWorksheetFromFile`
  * Read data from worksheets in an Excel file.

## Import Data from the Internet
***

* `httr::GET`: Get a URL.
* `jsonlite::read_json`: Read and write JSON.
* `readr::read_csv`
* `readr::read_csv2`
* `readr::read_delim`
* `readr::read_tsv`
* `rjson::fromJSON`: Convert JSON to R.
* `utils::download.file`
  * See example below.
* `utils::unzip`
  * See example below.

**Example:** `utils::download.file`

```{r, eval=FALSE}
download.file(
  "https://assets.datacamp.com/production/repositories/5028/datasets/a55843f83746968c7f118d82ed727db9c71e891f/snake_river_visits.rds",
  destfile = paste0(getwd(), "/Snake River Visits.rds"))

snake_river_visits <- readRDS(file.choose())
# or
path <- paste0(getwd(), "/Snake River Visits.rds")
snake_river_visits <- readRDS(path)
```

**Example:** Download `.Rdata` files

```{r, eval=FALSE}
# Example 1:
download.file(
  "https://assets.datacamp.com/production/repositories/236/datasets/7f714f993f1ad4c3d26412ae1e537ce6355b1b54/iris.RData", 
  destfile = "datacamp_iris_dataset.Rdata")

load("datacamp_iris_dataset.Rdata")

# Example 2:
download.file(
  "https://assets.datacamp.com/production/repositories/235/datasets/3b6fc2923b599058584b57d8c605c6bef454d273/CHIS2009_reduced_2.Rdata",
  destfile = "chis_2009.Rdata",
  # The documentation for `download.file` indicates that the function will
  # automatically include `mode = "wb"` for .Rdata files. That may have happened
  # in Example 1, but didn't happen in Example 2, which is why I've included it.
  mode = "wb")

load("chis_2009.Rdata")
```

**Example:** `utils::unzip`

```{r, eval=FALSE}
download.file(
  "https://assets.datacamp.com/production/repositories/1069/datasets/578834f5908e3b2fa575429a287586d1eaeb2e54/countries2.zip",
  destfile = "Data Sets/Countries",
  mode = "wb")

unzip("Data Sets/Countries", exdir = "Data Sets")
```

## Import from or Create Data in a Database
***

* `DBI::dbBind`
* `DBI::dbClearResult`
* `DBI::dbConnect`
* `DBI::dbDataType`
* `DBI::dbDisconnect`
* `DBI::dbExecute`
* `DBI::dbFetch`
* `DBI::dbGetQuery`
* `DBI::dbListTables`
* `DBI::dbReadTable`
* `DBI::dbSendQuery`
* `DBI::dbSendStatement`
* `dbplyr::tbl`

## Notes & References
***

* ["Data structures"](http://adv-r.had.co.nz/Data-structures.html) (Wickham, [*Advanced R*](http://adv-r.had.co.nz/))
* `{openintro}` contains data sets useful for practicing and teaching.
* ["Vectors"](https://r4ds.had.co.nz/vectors.html) (Grolemund & Wickham, [*R for Data Science*](https://r4ds.had.co.nz/))

<!--chapter:end:01-import.Rmd-->

# Tidy
***

"Tidying your data means storing it in a consistent form that matches the semantics of the dataset with the way it is stored. In brief, when your data is tidy, each column is a variable and each row is an observation. Tidying data is important because the consistent structure lets you focus your struggle on questions about the data, not fighting to get the data into the right form for different functions."  
- Garrett Grolemund & Hadley Wickham, *R for Data Science*
***

## Explore Raw Data
***

* `base::attr`
  * `attr(x, "names")` is the same as `base::names(x)`.
  * Use `attr(x, "names") <- value` to set attribute values.
* `base::attributes`
* `base::class`
* `base::colnames`
* `base::dim`
* `base::dimnames`
* `base::format`
* `base::length`
* `base::levels`
  * Levels default to alphabetical order, so be careful when renaming them (i.e., don't accidentally set the "F" level equal to "Male" rather than "Female").
* `base::mode`
* `base::names`
* `base::nchar`
* `base::order`
* `base::rownames`
* `base::summary`
* `base::typeof`
* `graphics::hist`
* `graphics::plot`
* `tibble::glimpse`
  * Similar to `utils::str`.
* `utils::head`
* `utils::str`
  * Similar to `tibble::glimpse`.
* `utils::tail`

## Tidy Data
***

* `base::duplicated`
  * See also `data.table::duplicated`
* `base::unique`
  * See also `data.table::unique`
* `data.table::anyDuplicated`
* `data.table::duplicated`
  * See also `base::duplicated`
* `data.table::unique`
  * See also `base::unique`
* `janitor::clean_names`
* `janitor::get_dupes`
  * Prefer over `base::duplicated` and `data.table::duplicated`.
* `janitor::remove_empty`
* `purrr::transpose`
  * Turn a list-of-lists inside-out.
* `splitstackshape::cSplit`
  * See also `tidyr::separate`
* `tibble::rowid_to_column`
  * Add a column of sequential row IDs.
  * Useful when a separate ID is required to manipulate rows or columns, such as when using `tidyr::pivot_longer`
* `tibble::rownames`
* `tidyr::nest`
  * Nest repeated values in a list-variable.
  * Helpful when separating a data frame in preparation to model the data for each grouping.
* `tidyr::replace_na`
* `tidyr::separate`
* `tidyr::pivot_longer`
* `tidyr::pivot_wider`
* `tidyr::unite`
* `tidyr::unnest`

## Prepare Data for Analysis
***

### Coerce

* `base::as.array`
* `base::as.data.frame`
  * See also`tibble::as_tibble`.
* `base::as.Date`
* `base::as.factor`
* `base::as.list`
* `base::as.matrix`
* `base::as.POSIX*`
* `base::factor`
  * Rearrange the order of factors by using the `levels` argument. For example, rearrange the order of "Bad, "Good," and "Neutral" using `levels = c("Bad", "Neutral", "Good").
* `base::is.na`
  * Use `is.na <-` to set elements to `NA`.
* `base::unclass`
* `methods::as`
* `tibble::as_tibble`
  * Preferable to `base::as.data.frame`.
* `tibble::enframe`
  * Preferable to using `tibble::as_tibble` to coerce a vector to a data frame.

### Dates and Datetimes

* `anytime::anytime`: Parse POSIXct or Date objects from input data.
* `base::as.Date()`
* `base::OlsonNames()` 
  * Displays available time zones.
* `base::as.POSIX*()`
* `base::as.POSIXct()`
* `base::strptime()`
  * Date-time conversion to and from character.
* `base::Sys.timezone()`
  * See also `base::OlsonNames`.
* `fasttime::fastPOSIXct`
  * Convert strings into `POSIXct` object (string must be in year, month, day, hour, minute, second format.)
* `hms:as.hms`
* `hms::hms`
  * Store time-of-day values as `hms` class.
* `hms::is.hms`
* `lubridate::as_date`
* `lubridate::fast_strptime`
  * Fast C parser of numeric formats only that accepts explicit format arguments (just as `base::strptime`).
  * Note that the format argument must match the input exactly, including any non-white space characters (such as "T" and "Z").
* `lubridate::make_date`
  * Create dates from numeric representations.
* `lubridate::make_datetime`
  * Create date-times from numeric representations.
* `lubridate::parse_date_time`
  * This function can be slow, because it is designed to be forgiving and flexible. If the dates you are working with are in a consistent format (ideally ISO 8601), use one of the following: `fasttime::fastPOSIXct` or `lubridate::parse_date_time2`.
* `lubridate::parse_date_time2`
  * Fast C parser of numeric orders.
* `lubridate::ymd`: Parse dates with year, month, and day components.
  * Related formats: `ydm`, `mdy`, `myd`, `dmy`, `dym`, `yq`.
* `lubridate::ymd_hms`: Parse date-times with year, month, day, hour, minute, and second components.
  * Related formats: `ymd_hm`, `ymd_h`, `dmy_hms`, `dmy_hm`, `dmy_h`, `mdy_hms`, `mdy_hm`, `mdy_h`, `ydm_hms`, `ydm_hm`, `ydm_h`.
        
### Filter

* `dplyr::filter`
* `purrr::keep`
* `stats::na.omit`

### Strings

* `base::cat`
  * Concatenate and print. 
* `base::chartr`
  * Change certain characters.
* `base::gregexpr`
* `base::grep`
* `base::grepl`
* `base::gsub`
* `base::regexec`
* `base::regexpr`
* `base::sub`
* `base::tolower`
  * See also `stringr::str_to_lower`.
* `base::toupper`: Convert to uppercase.
  * See also `stringr::str_to_upper`.
* `qdap::check_spelling`
* `qdapDictionaries::DICTIONARY`
  * Nettalk Corpus syllable data set.
* `qdapDictionaries::GradyAugmented`
  * Augmented list of Grady Ward's *English Words* and Mark Kantrowitz's *Names* List.
  * Mark Kantrowitz's *Names* list is available in full [here](http://www.cs.cmu.edu/afs/cs/project/ai-repository/ai/areas/nlp/corpora/names/).
* `stringr::str_detect`
  * Control the `pattern` argument options with `regex` (e.g., `str_detect(x, regex(pattern, ignore_case = TRUE))`.
* `stringr::str_remove`
* `stringr::str_to_lower`
* `stringr::str_to_title`
* `stringr::str_trim`
* `stringr::str_to_upper`

### Test

* `base::all()`
* `base::any`
  * Use `any(is.na(data.frame))` to determine if there are any NA values in a data frame.
* `base::anyNA`
  * Possibly faster implementation of `any(is.na(x))`.
* `base::exists`
* `base::is.array`
* `base::is.data.frame`
* `base::is.matrix`
* `base::is.vector`
* `base::setequal`
  * Check two vectors for equality.
* `base::sum`
  * Test whether all elements of a vector do or do not meet a certain condition, use as follows: `sum(email$num_char < 0)`.
* `purrr::every`
* `stats::complete.cases`
  * Find complete cases (i.e., rows without `NA` values).
* `tibble::is_tibble`

<!--chapter:end:02-tidy.Rmd-->

# Transform
***

"Transformation includes narrowing in on observations of interest (like all people in one city, or all data from the last year), creating new variables that are functions of existing variables (like computing velocity from speed and time), and calculating a set of summary statistics (like counts or means)."  
- Garrett Grolemund & Hadley Wickham, *R for Data Science*
***

## Arithmetic & Summary Statistics
***

* `assertive::is_divisible_by`
* `base::abs`
* `base::colMeans`
* `base::colSums`
* `base::diff`
* `base::IQR`
* `base::max`
* `base::mean`
  * Note that `base::mean` can be used to calculate a percentage when used in `dplyr::summarize`, as follows:
  
```{r, eval=FALSE}
by_country <- 
  votes %>% 
  group_by(country) %>% 
  summarize(total = n(),
            percent_yes = mean(vote == 1))
```
  
* `base::median`
* `base::min`
* Mode: Use `table` to view the mode of a data set.
* Operators:
  * Arithmetic Operators: `+`, `-`, `*`, `/`, `%%`
  * Comparison Operators: `<`, `>`, `<=`, `>=`, `==`, `!=`.
    * Use `identical` and `all.equal` rather than `==` and `!=` in tests where a single `TRUE` or `FALSE` is required (such as `if` expressions).
* `base::range`
  * Use `diff(range())` to get the range as a measure of variability.
* `base::round`
* `base::rowMeans`
* `base::rowSums`
* `base::sd`
* `base::signif`
  * Round values to a specified number of significant digits.
* `base::sqrt`
* `base::sum`
* `base::summary`
* `base::var`
* `dplyr::count`
* `dplyr::group_by`
* `dplyr::n`
  * Must be used within `summarise`, `mutate`, or `filter`.
* `dplyr::n_distinct`
* `dplyr::near`
  * Compare two numeric vectors (safer than using `==`).
* `dplyr::summarize`
  * Use `mean(variable == value)` to get a percentage (see the above example for `mean`.)
* `dplyr::tally`
* `dplyr::top_n`
* `magrittr::extract`
  * `x %>% extract(y)` is equivalent to `x[y]`. 
* `magrittr::multiply_by`
  * `x %>% multiply_by(y)` is equivalent to `x * y`.
* `magrittr::raise_to_power`
  * `x %>% raise_to_power(y)` is equivalent to `x^y`.
* `stats::aggregate`
  * Compute summary statistics of data subsets.
* `stats::cor`
* `stats::cov`
* `stats::lag`
* `stats::rnorm`
* `stats::var`

***
## Create New Variables or Modify Existing Ones

`countrycode`

* `countrycode()`: Convert country codes into country names.

`dplyr`

* `mutate()`: Add new variables.
    + `mutate()` can also be used to modify existing variables. To change the case of a character variable, for example, do something like:
    
```{r, eval=FALSE}
df <-
  df %>%
  mutate(var_name = str_to_lower(var_name))
```

    + Child function: `transmute()` (drops existing variables).
* `recode()`: Recode values (the numeric alternative to using `if_else` or `case_when()`).
    
***
## Dates and Datetimes

* `base::date`
  * Get the current system date and time.
* `base::difftime`
  * Time intervals and differences.
  * `base::difftime` is the function behind the `-` operator when used with dates and datetimes (e.g., `time_1 - time_2` is equivalent to `difftime(time_1, time_2)`). The advantage of using  `difftime` over `-`, however, is the `units` argument because it allows you to specify the unit of time in which the difference is calculated.
* `base::months`
  * Extract the month names.
* `base::quarters`
  * Extract the calendar quarters.
* `base::Sys.Date`
  * Get the current date in the current time zone.
* `base::Sys.time`
  * Get the absolute date-time value (which can be converted to various time zones and may return different days).
* `base::weekdays`
  * Extract weekday names.
* `lubridate::ceiling_date`
* `lubridate::date`
  * Get or set the date component of a date-time.
* `lubridate::day`
  * Get or set the day component of a datetime.
* `lubridate::floor_date`
* `lubridate::month`
  * Get or set the month component of a datetime.
* `lubridate::now`
  * The current time (as a POSIXct object).
* `lubridate::quarter`
  * Get or set the fiscal quarter or semester component of a datetime.
* `lubridate::round_date`
* Time spans: Duration
  * Use when you are interested in seconds elapsed.
  * `lubridate::ddays`
  * `lubridate::dhours`
  * `lubridate::dminutes`
  * `lubridate::dseconds`
  * `lubridate::dweeks`
  * `lubridate::dyears`
* Time spans: Interval
  * Use when you have a start and end.
  * `%--%`
  * `int_aligns`
  * `int_diff`
  * `int_end`
  * `int_flip`
  * `int_length`
  * `int_overlaps`
  * `int_shift`
  * `int_standardize`
  * `int_start`
  * `interval`
  * `is.interval`
* Time spans: Period
  * Use when you are interested in human units.
  * `lubridate::days`
  * `lubridate::hours`
  * `lubridate::minutes`
  * `lubridate::months`
  * `lubridate::seconds`
  * `lubridate::weeks`
  * `lubridate::years` 
* Time zones:
  * `lubridate::force_tz`
    * Change the time zone without changing the clock time.
  * `lubridate::tz`
    * Extract the time zone from a datetime.
  * `lubridate::with_tz`
    * View the same instant in a different time zone.
* `lubridate::today`
* `lubridate::%m+%` & `%m-%`
  * Add and subtract months to a date without exceeding the last day of the new month.
* `lubridate::%within%`
  * Test whether a date or interval falls within an interval.
* `lubridate::year`
  * Get or set the year component of a datetime.

***
## Factors

`forcats`

* `fct_drop()`: Drop levels.
* `fct_reorder()`: Reorder levels, based on the value of another variable.
* `fct_rev()`: Reverse levels.

`stats`

* `reorder()`: Reorder levels of a factor.
    + Useful within the `aes()` argument in a `ggplot()` call.

## Merge or Append Data

`base`

* `append()`: Add elements to a vector.
* `cbind()`: Combine objects by column.
* `intersect()`: Combine data shared in common between two datasets.
    + Similar to `dplyr::semi_join()`.
* `merge()`: Merge two data frames.
    + `dplyr::join` functions are an alternative to `merge()`.
* `rbind()`: Combine objects by row.
* `setdiff()`: Find the difference between two vectors.
    + Similar to `dplyr::anti_join()`.
* `union()`: Combine two datasets without duplicating values.

`dplyr`

* `bind()`: Bind multiple data frames by row and column.
    + Child functions: bind_rows(), bind_cols(), combine().
* Join Functions: Join two tables.
    + Filtering Joins:
        + anti_join(): Return all rows from `x` where there are not matching values in `y`, keeping just columns from `x`.
        + semi_join(): Return all rows from `x` where there are matching values in `y`, keeping just columns from `x`. A semi join differs from an inner join because an inner join will return one row of `x` for each matching row of `y`, where a semi join will never duplicate rows of `x`.
    + Mutating Joins:
        + full_join(): Return all rows and all columns from both `x` and `y`. Where there are not matching values, returns `NA` for the one missing.
        + inner_join(): Return all rows from `x` where there are matching values in `y`, and all columns from `x` and `y`. If there are multiple matches between `x` and `y`, all combination of the matches are returned.
        + left_join(): Return all rows from `x`, and all columns from `x` and `y`. Rows in `x` with no match in `y` will have `NA` values in the new columns. If there are multiple matches between `x` and `y`, all combinations of the matches are returned.
        + right_join(): Return all rows from `y`, and all columns from `x` and `y`. Rows in `x` with no match in `y` will have `NA` values in the new columns. If there are multiple matches between `y` and `x`, all combinations of the matches are returned.

`tibble`

* `add_column()`: Add columns to a data frame.
* `add_row()`: Add rows to a data frame.

***
## Narrow in on Observations of Interest

`{base}`

* `droplevels`: Drop unused levels from factors.
  * This function will keep levels that have even 1 or 2 counts. If you want to remove levels with low counts from a data set in order to simplify your analysis, first `filter` out those rows and then use `droplevels`.
* `order`: Arrange in ascending or descending order.
  * Alternative: `dplyr::arrange`.
* `prop.table`: Express table entries as proportions of the marginal table.
  * The input is a table produced by `table`.
  * As these are proportions of the whole, `sum(prop.table(table_name))` = 1.
  * Specify conditional proportions on rows or columns by using the `margin` argument.
* `table`: Build a table of the counts at each combination of factor levels.
  * Use `prop.table` to see the table entries expressed as proportions.

`{dplyr}`

* `arrange`: Arrange rows by variable, in ascending order.
    + Related functions: `arrange_all`, `arrange_at`, `arrange_if`.
* `distinct`: Select distinct rows.
* `filter`: Return rows with matching conditions.
  * Use `%in%` when using multiple `|` conditions. The following two commands, for example, are equivalent:
```{r, eval=FALSE}
ilo_data %>% 
  filter(country %in% c("Sweden", "Switzerland"))

ilo_data %>% 
  filter(country == "Sweden" | country == "Switzerland")
```
    
* `renam`: Rename variables by name (a modification of `select`).
* `sample_n`: Sample n rows from a table.
* `select`: Select/rename variables. 
  * Helper functions include: `starts_wit`, `ends_with`, `contains`, `matches`, `num_range`, and `one_of`. See `?dplyr::select_helpers`.
  * Helper functions take strings (e.g., `contains("work")` rather than `contains(work)`).
* `slice`: Choose rows by position.
* `transmute`: Create or transform variables.
  * Like a combination of `select` and `mutate`.

***
## Test

`base`

* `identical()`: Test objects for exact equality.
* `match()`: Value matching.
    + `%in%` is the more intuitive binary operator.
* `setequal()`: Check two vectors for equality.
* `which()`: Determine which indices are `TRUE`.
    + This function is often unnecessary, according to the DataCamp article ["Five Tips to Improve Your R Code"](https://www.datacamp.com/community/tutorials/five-tips-r-code-improve).
* `which.min()`/`which.max()`: Where is the `min()`/`max()` or first `TRUE`/`FALSE`?
* `%in%`: See `match()`.

<!--chapter:end:03-transform.Rmd-->

# Visualize
***

"Visualisation is a fundamentally human activity. A good visualisation will show you things that you did not expect, or raise new questions about the data. A good visualisation might also hint that you're asking the wrong question, or you need to collect different data. Visualisations can surprise you, but don't scale particularly well because they require a human to interpret them."  
- Garrett Grolemund & Hadley Wickham, *R for Data Science*
***

## Interfaces
***

* `shiny`

## Graphs, Charts, and Plots
***

* `base::abline`
  * Add straight lines to a plot.
* `base::plot`
* `base::points`
  * Add points to a plot.
* `diagram`
* `DiagrammeR`
* `ggbeeswarm::geom_beeswarm`
* `ggplot2::aes`
  * Arguments include `color`, `fill`, `size`, `labels`, `alpha`, `shape` (1-20 accept color attributes and 21-25 accept color and fill attributes), `linewidth`, `linetype`, and `group`.
  * Use `?pch` to see options for `shape`.
  * Note that `aes` can be called within `geom_*` rather than prior to.
  * The default shape for points does not have a `fill` attribute, which means that mapping a categorical variable onto `fill` won't result in multiple colors.
  * Helper functions to include in the call when needing to modify the data include: `stats::reorder`.
* `ggplot2::coord_cartesian`
  * Zoom a plot in or out without changing the underlying data.
* `ggplot2::coord_flip`
  * Flip the x and y axes.
* `ggplot2::coord_polar`
  * Use to convert a stacked bar chart to a pie chart.
* `ggplot2::element_blank`
* `ggplot2::element_line`
* `ggplot2::element_rect`
* `ggplot2::element_text`
* `ggplot2::facet_grid`
  * Lay out panels in a grid.
  * When using a tilde function, the variable on the left specifies the rows and the variable on the right specifies the columns (when faceting in only one direction; i.e., only rows or only columns). Use `.` to specify nothing for the unused direction (e.g. `facet_grid(cyl ~ .)`).
* `ggplot2::facet_wrap`
  * Wrap a 1D ribbon of panels into 2D (observe a variable, conditional on another variable).
* `ggplot2::geom_abline`
  * Add reference lines to a plot.
* `ggplot2::geom_bar`
  * Create a bar chart, where the height of the bar is proportional to the number of cases in each group.
* `ggplot2::geom_boxplot`
* `ggplot2::geom_col`
  * Create a bar chart, where the height of the bar represents values in the data.
* `ggplot2::geom_density`
  * Create a kernal density estimate (a smoothed version of a histogram).
  *  Consider using `ggplot2::geom_rug` with `ggplot2::geom_density` in order to be transparent about smoothed data.
* `ggplot2::geom_dotplot`
  * Create a histogram out of dots.
* `ggplot2::geom_errorbar`
  * Shortcut for `geom_bar(stat = "identity")`.
* `ggplot2::geom_histogram`
  *  Use `y = stat(density)` to rescale the y-axis from counts to a probability estimate.
* `ggplot2::geom_hline`
*`ggplot2::geom_jitter`
  * Useful with `ggplot2::geom_boxplot`.
* `ggplot2::geom_path`
  * Connect observations in the order in which they appear.
* `ggplot2::geom_point`
* `ggplot2::geom_rug`
  * Useful with `ggplot2::geom_density`.
* `ggplot2::geom_smooth`
  * Smoothed conditional means; aids the eye in seeing patterns in the presence of overplotting.
* `ggplot2::geom_text`
* `ggplot2::geom_violin`
* `ggplot2::geom_vline`
* `ggplot2::ggplot`
* `ggplot2::ggtitle`
* `ggplot2::group`
  * Used within `aes`. 
  * Note that `group` is usually unnecessary when specifying `color`, `shape`, `fill`, or `linetype` within `aes`, or when using facets. See ["Aesthetics: grouping"](https://ggplot2.tidyverse.org/reference/aes_group_order.html) for more information.
* `ggplot2::labs`
  * Modify axis, legend, and plot labels.
* `ggplot2::position_identity`
  * Don't adjust position.
* `ggplot2::position_dodge`
  * Dodge overlapping objects side-to-side.
* `ggplot2::position_nudge`
  * Nudge points a fixed distance.
* `ggplot2::position_stack`
  * Stack overlapping objects on top of each other, as counts.
* `ggplot2::position_fill`
  * Stack overlapping objects on top of each other, as densities.
* `ggplot2::position_jitter`
* `ggplot2::position_jitterdodge`
* `ggplot2::scale_x_continuous`
* `ggplot2::scale_y_*`
* `ggplot2::scale_color_manual`
* `ggplot2::scale_fill_*`
* `ggplot2::scale_shape_*`
* `ggplot2::scale_size`
* `ggplot2::scale_linetype_*`
* `ggplot2::stat_bin`
* `ggplot2::stat_bin2d`
* `ggplot2::stat_bindot`
* `ggplot2::stat_binhex`
* `ggplot2::stat_boxplot`
* `ggplot2::stat_contour`
* `ggplot2::stat_quantile`
* `ggplot2::stat_smooth`
* `ggplot2::stat_sum`
* `ggplot2::theme`
  * See example for `grid::unit`.
* `ggplot2::theme_get`
* `ggplot2::theme_set`
* `ggplot2::theme_update`
  * "When you call 'theme_update' and assign it to an object (e.g., called `old`), that object stores the *current* default theme, and the arguments *update* the default theme. If you want to restore the previous default theme, you can get it back by using `theme_update` again." -DataCamp
* `ggplot2::xlab`
* `ggplot2::xlim`
* `ggplot2::ylab`
* `ggridges::geom_density_ridges`
  * Create a ridgeline plot.
* `graphics::boxplot`
  * Create a box-and-whisker plot.
* `graphics::hist`
* `graphics::stripchart`
  * One dimensional scatter plots.
  * Prefer over `ggplot` when creating one-dimensional plots.
* `grDevices::colorRamp`
  * Color interpolation.
* `grDevices::colorRampPalette`
  * Color interpolation.
* `grid::unit`
  * Create a `unit` object, to specify locations and dimensions in a coordinate system.
  * See example below.
* `RColorBrewer::brewer.pal`
  * Make the ColorBrewer color palettes available as R palattes.
  
**Example: `grid::unit`

```{r, eval=FALSE}
# Increase spacing between facets in a `ggplot` object:

plot + 
  theme(
    panel.spacing.x = unit(2, "cm"),
    panel.margin = unit(c(1, 2, 1, 1), "cm"))
```

## Graphs, Charts, and Plots - Statistics
***

* `ggplot2::mean_cl_boot`
* `ggplot2::mean_cl_normal`
* `ggplot2::sdl`
* `ggplot2::median_hilow`
* `ggplot2::stat_function`
  * Compute `y` values from a function of `x` values.
* `ggplot2::qq`
  * Perfrm calculations for a quantile-quantile plot.
* `ggplot2::qq_line`
* `ggplot2::summary`
  * Summarize `y` values at distinct `x` values.

<!--chapter:end:04-visualize.Rmd-->

# Model
***

"Models are complementary tools to visualisation. Once you have made your questions sufficiently precise, you can use a model to answer them. Models are a fundamentally mathematical or computational tool, so they generally scale well. ... But every model makes assumptions, and by its very nature a model cannot question its own assumptions. That means a model cannot fundamentally surprise you. 
- Garrett Grolemund & Hadley Wickham, *R for Data Science*"
***

## General

* `base::sample`
* `dplyr::sample_n`
* `stats::coef`
  * Extract model coefficients.
* `stats::cor`
* `stats::cov`
* `stats::cov2cor`
  * Scale a covariance matrix into a correlation matrix.
* `stats::var`

***
## Regression

* `broom::augment`
  * Augment data with information from an object.
* `broom::glance`
  * Construct a single row summary of a model, fit, or other object.
* `broom::tidy`
  * Turn an object into a tidy tibble.
* `{mgcv::gam`
  * Generalized additive models (GAMs) with integrated smoothness estimation.
* `mixtools::`
* `stats::df.residual`
  * Get the residual degrees of freedom.
* `stats::lm`
  * Fit linear models.
* `stats::p.adjust`
  * Adjust p-values for multiple comparisons.
* `stats::predict`
  * Model predictions.
* `stats::residuals`
  * Extract model residuals.
* `tidyr::nest`
  * Nest repeated values in a list-variable.
  * Helpful when separating a data frame in preparation to model the data for each grouping.

***
## Simulation & Prediction

* `base::set.seed`
* `stats::predict`
  * Model prediction.
  * Use with `lm` class objects and new data to predict new values (e.g., `predict(model, newdata)`).
* `stats::rnorm`

## Strings
***

* `base::agrep`
* `fuzzyjoin::stringdist_join`
* `fuzzyjoin::stringdist_anti_join`
* `fuzzyjoin::stringdist_full_join`
* `fuzzyjoin::stringdist_inner_join`
* `fuzzyjoin::stringdist_left_join` 
* `fuzzyjoin::stringdist_right_join`
* `fuzzyjoin::stringdist_semi_join`
* `fuzzywuzzyR::FuzzMatcher`

<!--chapter:end:05-model.Rmd-->

# Communicate
***

"The last step of data science is communication, an absolutely critical part of any data analysis project. It doesn't matter how well your models and visualisation have led you to understand the data unless you can also communicate your results to others."  
- Garrett Grolemund & Hadley Wickham, *R for Data Science*
***

## Export
***

* `base::file.path`
* `base::print`
  * Use the `include.rownames = FALSE` argument to remove row numbers (or names) from the output.
* `base::save`
* `base::saveRDS`
  * See ["A better way of saving and loading objects in R"](https://www.fromthebottomoftheheap.net/2012/04/01/saving-and-loading-r-objects/) to understand the differences between `save` and `saveRDS`.
* `readr::write_csv`
* `readr::write_delim`
  * About twice as fast as `utils::write.csv`, and never writes row names.
* `readr::write_excel_csv`
* `readr::write_tsv`
* `utils::write.csv`
* `utils::write.csv2`
* `utils::write.table`
  * Prefer `readr::write_delim` to `utils::write.table`.
* `XLConnect::loadWorkbook`

## Format Output
***

* `base::format`
* `DT::datatable`
  * Use the `options = list(dom = 't')` argument to remove the search bar.
  * Use the `options = list(columnDefs = list(list(className = "db-center", targets = (column_position))))` argument to center column values (where `0` indicates the first column).
* `kableExtra::kable_styling`
* `kableExtra::add_footnote`
* `knitr::kable`
* `lubridate::stamp`
  * Format dates and times based on human-friendly templates.
* `scales::dollar`

## Graphics
**

* `knitr::include_graphics`
  * Embed external images in 'knitr' documents.
  * Preferable to the `![alt text or image title](path/to/image)` Markdown syntax for embedding an external image, as `include_graphics` offers more control over the attributes of the image.

## Plots
***

* `ggplot2::element_*`: Specify the display of how non-data components of a plot are drawn.
* `ggplot2::labs`: Modify axis, legend, and plot labels.
    + Child functions: `xlab`, `ylab`, `ggtitle`
* `ggplot2::theme`: Customize the non-data components of a plot.
  * See example below.
  * See `?theme_classic` for a list of predefined themes.

**Example: `ggplot2::theme`**

```{r, eval=FALSE}
ggplot(plot_data_2006) +
  geom_histogram(aes(x = working_hours)) +
  labs(
    x = "Working hours per week",
    y = "Number of countries") +
  theme(
    text = element_text(family = "Bookman", color = "gray25"))
```
    
## RMarkdown
***

* See [RMarkdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)
* See [Mozzila Developer Network](https://developer.mozilla.org/en-US/docs/Web/CSS/CSS_Selectors) for CSS help.

<!--chapter:end:06-communicate.Rmd-->

# Program
***

"Surrounding [the tools for importing, tidying, transforming, visualising, modeling, and communicating data] is programming. Programming is a cross-cutting tool that you use in every part of a project. You don't need to be an expert programmer to be a data scientist, but learning more about programming pays off because becoming a better programmer allows you to automate common tasks, and solve new problems with greater ease."  
- Garrett Grolemund & Hadley Wickham, *R for Data Science*
***

## Characters

References:
* `?Quotes`

## Conditionals & Control Flows
***

`base`

* Control (access documentation using "?Control")
    + `if (cond) expr`
        + The key difference between `if (cond) expr` and `ifelse` is that `if (cond) expr` will evaluate only the first element of an object with `length > 1`. See the documentation for each function and the ateucher's GitHub example titled ["R: if vs ifelse"](https://gist.github.com/ateucher/c7359f566eded9fcd4a255f4cbd4fe67) to learn more.
    + `for (var in seq) expr`
    + `while (cond) expr`
    + `repeat expr`
    + `break`
    + `next`
* `identical()`: Test objects for exact equality.
    + Use `identical()` rather than `==` and `!=` in `if` and `while` statements to test for equality.
* 'ifelse()`: Conditional element selection.
    + `dplyr::if_else()` is more strict by checking the object type.
* `stop()`: Stop execution of the expression and execute an error action.
    + Useful in combination with an `if` statement when you want to generate helpful error messages.
* `stopifnot()`: Ensure the truth of an R expression.
    + Prefer `base::stop()` to `base::stopifnot()`.

`dplyr`

* `case_when()`: A general vectorized `if`.
* `if_else()`: Vectorized `if`.

***
## Environment and Workspace

* `base::baseenve`
  * The environment of the `base` package; it's enclosing environment ("parent environment") is the empty environment.
* `base::dir`
  * List the files in a directory.
* `base::dir.create`
  * Create a file path in the computer's file system.
* `base::emptyenv`
  * The empty environment, which is the ancestor of all environments and the only environment without an enclosing environment.
* `base::environment`
  * The current environment.
* `base::environmentName`
* `base::exists`
  * Remember that `R` will look for an object in parent environments until it reaches the empty environment, so use `inherits = FALSE` to limit the search to only the current environment.
* `base::getOption`
  * Set and examine global options.
* `base::getwd`
  * Get the working directory.
* `base::globalenv`
  * The environment in which you normally work, it's enclosing environment is the last package attached with `library` or `require`.
* `base::history`
  * Display the previous 25 commands.
* `base::install.packages`
  * Install packages from repositories or local files.
* `base::library`
  * Load and attach packages, returning an error if the packages does not exist.
* `base::list.files`
  * List the files in a directory/folder.
* `base::loadedNamespaces`
  * Return the loaded name spaces.
* `base::loadhistory`
  * Recall command history.
* `base::ls`
  * List objects in the specified environment.
* `base::list2env`
  * From a list, build or add to an environment.
* `base::new.env`
  * Create a new environment.
* `base::options`
  * Set and examine global options.
* `base::parent.env`
  * Return the enclosing environment of the environment listed as an argument.
  * `base::parent.env` returns information that can be unhelpful, so use with `base::environmentName`, as follows: `parent.env(environment_name) %>% environmentName`.
* `base::q`
  * Terminate an R session.
* `base::R.version`
  * Version information.
* `base::R.version.string`
  * Version information. 
  * Same call as `R.version$version.string`.
* `base::require`
  * Load and attach packages, returning `FALSE` if the package does not exist.
* `base::rm`
  * Remove objects from a specified environment.
* `base::savehistory`
  * Save command history (default value is ".Rhistory").
* `base::save.image`
  * Save the current workspace.
* `base::search`
  * Return a list of attached packages and R objects.
* `base::searchpaths`
  * Return the path to attached packages.
* `base::setwd`
  * Set the working directory file path.
  * When using Windows, use "/" instead of "\".
* `base::Sys.getenv`
  * See also `Sys.setenv`.
* `base::Sys.info`
  * Extract system and user information.
  * Example: `Sys.info()[c("sysname", "release")]`.
* `base::Sys.setenv`
  * See also `Sys.setenv`.
* `gdata::object.size`
  * Report the space allocated for an object.
  * See also `utils::object.size`.
* `installr::updateR`
  * Check for the latest R version; downloads and installs new R versions.
* `pryr::where`: Find where a name is defined.
* `utils::ls.str`: List objects and  their structure.
* `utils::object.size`: Report the space allocated for an object.
  * See also `gdata::object.size`.
* `utils::sessionInfo`: Collect information about the current R session.
    
References:

* ["Environments"](http://adv-r.had.co.nz/Environments.html#environments) (Hadley Wickham, [*Advanced R*](http://adv-r.had.co.nz/))

***
## Evaluation (Standard and Non-standard)

`base`

* `cat()`: Concatenate and print.
* `print(): Print the argument to the Console.
    + A shortcut to `print()` is to place the code you want printed inside parentheses.
* `quote()`: Return the argument, unevaluated.
* `writeLines()`: Display quotes and backslashes as they would be read, rather than as R stores them (i.e., see the raw contents of the string, as the `print()` representation is not the same as the string itself).

`rlang`

* Quosures
    + `enquo()`, `new_quosure()`, `quo()`.
    
References:

* ["Non-standard evaluation"](http://adv-r.had.co.nz/Computing-on-the-language.html) (Hadley Wickham, [*Advanced R*](http://adv-r.had.co.nz/))
* ["Non-standard evaluation"](https://cran.r-project.org/web/packages/lazyeval/vignettes/lazyeval.html) (Hadley Wickham, `lazyeval` package vignette)
* ["Programming with dplyr"](https://dplyr.tidyverse.org/articles/programming.html) (dplyr.tidyverse.org)

***
## Functionals

`{base}`

* Apply Functions
  * `apply`: Apply functions over array margins.
  * `lapply`: Apply a function over a list or vector.
  * `sapply`: Apply a function over a list or vector and return a vector or matrix.
  * `vapply`: A safer version of `sapply`, as it requires the output type to be predetermined.
  * `mapply`: Apply a function to multiple list or vector arguments.
  * `rapply`: Recursively apply a function to a list. 
  * `tapply`: Apply a function over a ragged array.

`{purrr}`

* `map`: Apply a function to each element of a vector.
  * `tidyr::unnest` is useful in changing the list-column output of `map` into rows.
* `map2`: Map over multiple inputs simultaneously.
* `map_if`: Apply a function to elements of that match a condition.
* `possibly`: Usese a default value whenever an error occurs.
* `quietly`: Capture side effects in a list with components `result`, `output`, `messages`, and `warnings`.
* `safely`: Capture side effects in a list with components `result` and `error`.
* `transpose`: Transpose a list (turn a list-of-lists inside-out).

***
## Functions

`{assertive}`

* `assert_*`: Check whether the input is `*` (e..g, `assert_is_numeric`) and throw and error if the input does not meet the condition.
* `coerce_to`: Coerce the input to a different class, with a warning.
* `is_*`: Checks whether the input matches the condition specified by `*` (e.g., `is_non_positive`).
* `use_first`: Use only the first element of a vector.

`{base}`

* `do.call`: Execute a function call from a name or a function and a list of arguments to be passed to the function.
* `invisible`: Return a (temporarily) invisible copy of an object.
* `match.arg`: Argument verification.
  * Useful when matching a character argument specified in the function signature. For example, 

```{r}
args(prop.test)

# The body of `prop.test` contains the following line of code:
# `alternative <- match.arg(alternative), which reassigns it to the selected
# character vector.
```
  
* `message`: Generage a diagnostic message.
  * Preferable to generating a message using `cat`.
* `return`: Return a value from a function.
  * Useful in `if` statements where one condition is simple and the other is complex (see section 19.6.1 "Explicit return statements" in Hadley Wickham's [*R for Data Science*](https://r4ds.had.co.nz/functions.html).
* `setNames`: Set the names in an object.
  * Useful in function writing; see documentation.
* `stop`: Stop execution of the expression and execute an error action.
  * Useful in combination with an `if` statement when you want to generate helpful error messages.
* `stopifnot`: Ensure the truth of an R expression.
  * See section 19.5.2 "Checking values" in Hadley Wickham's [*R for Data Science*](https://r4ds.had.co.nz/functions.html) for a discussion of `stop` versus `stopifnot`.
  * Consider functions from `{assertive}` as an alternative to `stopifnot` and `stop`.
* `unlist`: Flatten lists.
  * Useful when using `purrr`'s `map` functions, which return objects as type `list`.

`{zeallot}`

* `%<-%`: Multiple assignment operator.
  * Useful when you want to return multiple output from a function. For example:
  
```{r, eval=FALSE}
session <- function() {
  
  list(
    r_version = R.version.string,
    operating_system = Sys.info()[c("sysname", "release")],
    loaded_pkgs = loadedNamespaces()
  )
}

c(vrsn, os, pkgs) %<-% session() 
```

***
## Learn About an Object

`base`

* `args()`: Display the argument names and default values of a function.
* `attributes()`: View or assign an objects attributes (e.g., `class()`, `dim()`, `dimnames()`, `names()`, `row.names()`). 
* `body()`: Get or set the body of a function.
* `colnames()`: Retrieve or set column names.
* `dim()`: Retrieve or set the dimnames of an object.
* `dimnames()`: Retrieve or set the dimension names of an object.
* `formals()`: Get or set the formal arguments of a function.
* `help()`: Get the topic documentation.
* `help.search()`: Search the help system for documentation matching a given character string.
* `vignette()`: View a specified package vignette.
* ?object_name
* ??object_name
* `rownames()`: Retrieve or set row names.

***
## Loops

`base`

* `seq()`: Sequence generation.
    + This functions makes `length()` unnecessary.
    + Child function: `seq_along()`
        + In `for` loops, safer than using `ncol()` or `nrow()`.

***
## Optimization

`microbenchmark`

* `microbenchmark()`: Sub-millisecond accurate timing of expression evaluations.
    + A more accurate replacement of `system.time(replicate(1000, expr))`.

***
## Pipes

`magrittr`

* `%<>%`: Compound assignment-pipe operator.
* `%>%`: Forward-pipe operator.
* `%$%`: Expositions-pipe operator.
* `add`: `+`, for pipes.
* `and`: `&`, for pipes.
* `extract`:`[`, for pipes.
  * See also `purrr::pluck`.
* `extract2`: `[[`, for pipes.
* `freduce`: Apply a list of functions sequentially.
* `is_in`: `%in%`, for pipes.
* `multiply_by`: `*`, for pipes.
* `or`: `|`, for pipes.
* `raise_to_power`: `^`, for pipes.
* `subtract`: `-`, for pipes.

***
## Popups

`svDialogs`

* `dlg_message`: Display a modal message box (works in Windows, MacOS, and Linux).

`tcltk`

* `tk_messageBox`: Display a generic message box using Tk (Windows-specific).

## Selecting & Subsetting

* `.$variable_name`
  * See example below.
* `.[["variable_name"]]`
  * See example below.
* `base::subset`
* `dplyr::first`
* `dplyr::last`
* `dplyr::nth`
* `dplyr::rename`
* `dplyr::select`
  * Helper functions: `contains`, `ends_with`, `matches`, `num_range`, `one_of`, `starts_with`.
* `magrittr::extract`
* `magrittr::extract2`
* `purrr:pluck`

**Example:** `.$variable_name`

```{r, eval=FALSE}
ui_summary_table <- 
  aws_vendors %>% 
  filter(str_detect(vendor_name, "UTAH INTERACTIVE")) %>% 
  .$vendor_id %>% 
  map(query_summary_table) %>% 
  bind_rows()
```

**Example:** `.[["variable_name"]]`

```{r, eval=FALSE}
  odbc_aws %>% 
    dbGetQuery(
      paste("
            SELECT id
            FROM batch
            WHERE entity_id = ", t_id, "
            AND status IN ('PROCESSED', 'PROCESSING')")) %>% 
    .[["id"]] %>% 
  as.double()
```

**References:**

* ["Indexing lists in #rstats. Inspired by Residence Inn"](https://twitter.com/hadleywickham/status/643381054758363136) (Hadley Wickham, Twitter, 14 September 2015)

***
## System Commands

* `base::system`
* `base::system2`
* `sys::exec`
* `sys::exec_r`

## Version Control

Git

* [Git](https://git-scm.com/)
* [*Pro Git*](https://git-scm.com/book/en/v2) by Scott Chacon and Ben Straub
* [*Git and GitHub*](http://r-pkgs.had.co.nz/git.html#git-learning) by Hadley Wickham
* [*Happy Git and GitHub for the useR*](http://happygitwithr.com/) by Jenny Bryan

* `git branch`: List, create, or delete branches.
    + `git branch -d <branch_name>`: Delete a local branch.
        + See ["Delete branch in RStudio pop-up"](https://community.rstudio.com/t/delete-branch-in-rstudio-pop-up/15465) for help removing branches in RStudio after removing them from Git.

`{packrat}`

* See note on the `packrat` package in the "Referenced Packages" section.
* `snapshot`: Capture and store the packages and versions in use.
* `restore`: Load the most recent snapshot to the project's private library.

<!--chapter:end:07-program.Rmd-->

# Referenced Packages
***
## A-D

[`anytime`](https://CRAN.R-project.org/package=anytime): Date converter.

[`assertive`](https://CRAN.R-project.org/package=assertive): Check functions to ensure code integrity.

[`base`](https://www.rdocumentation.org/packages/base/versions/3.5.1): Base R functions.

[`bookdown`](https://CRAN.R-project.org/package=bookdown): Author books and technical documents with R Markdown.
    + See Yihui Xie's [*bookdown: Authoring Books and Technical Documents with R Markdown*](https://bookdown.org/yihui/bookdown/).

[`broom`](https://CRAN.R-project.org/package=broom): Convert statistical analysis objects into tidy data frames.

[`car`](https://CRAN.R-project.org/package=car): Companion to applied regression.

[`chron`](https://CRAN.R-project.org/package=chron): Chronological objects which can handle dates and times.

[`countrycode`](https://CRAN.R-project.org/package=countrycode): Convert country names and country codes.

[`data.table`](https://CRAN.R-project.org/package=data.table): For large data.

[`DBI`](https://CRAN.R-project.org/package=DBI){#DBI}: Database interface.

* For MySQL documentation, see the [MySQL Reference Manual](https://dev.mysql.com/).
    
* Use with the `odbc` package.

[`diagram`](https://CRAN.R-project.org/package=diagram): Visualize simple graphs (networks); create plot flow diagrams.

[`DiagrammeR`](https://CRAN.R-project.org/package=DiagrammeR): Graph/network visualization.
    + DiagrammeR uses the [GraphViz](https://www.graphviz.org/) language.

[`dplyr`](https://CRAN.R-project.org/package=dplyr): Data manipulation.

* See also [dplyr.tidyverse.org](https://dplyr.tidyverse.org/).

***
## E-H

[`fasttime`](https://CRAN.R-project.org/package=fasttime): Fast utilit function for time parsing and conversion.

[`forcats`](https://CRAN.R-project.org/package=forcats): Tools for working with categorical variables.

[`fuzzyjoin`](https://CRAN.R-project.org/package=fuzzyjoin): Join tables together on inexact matching.

[`fuzzywuzzyR`](https://CRAN.R-project.org/package=fuzzywuzzyR): Fuzzy string matching.

[`gdata`](https://CRAN.R-project.org/package=gdata): Data manipulation.

[`ggbeeswarm`](https://CRAN.R-project.org/package=ggbeeswarm): Categorical scatter plots (violin and beeswarm).

[`ggplot2`](https://CRAN.R-project.org/package=ggplot2): Create elegant data visualizations.

[`ggridges`](https://CRAN.R-project.org/package=ggridges): Ridgeline plots in `ggplot2`.

[`graphics`](https://www.rdocumentation.org/packages/graphics/versions/3.5.1): R functions for base graphics.

[`grDevices`](https://www.rdocumentation.org/packages/grDevices/versions/3.5.2): Grahpics devices and support for base and grid graphics.

[`Hmisc`](https://CRAN.R-project.org/package=Hmisc): Harrell miscellaneous.

[`hms`](https://CRAN.R-project.org/package=hms): Times without dates.

[`httr`](https://CRAN.R-project.org/package=httr): Tools for working with HTTP.

***
## I-L

[`installr`](https://CRAN.R-project.org/package=installr): Install and update stuff (such as R, Rtools, Rstudio, Git).

[`janitor`](https://CRAN.R-project.org/package=janitor): Simple tools for examining and cleaning dirty data.

[`jsonlite`](https://CRAN.R-project.org/package=jsonlite): A robust, high performance JSON parser and generator.

[`knitr`](https://CRAN.R-project.org/package=knitr): Dynamic report generation in R using Literate Programming techniques.
    + See Yihui Xie's [*knitr*](http://yihui.name/knitr/).

[`lubridate`](https://CRAN.R-project.org/package=lubridate): Functions to work with date-times and time-spans.

* `lubridate` uses character formatting similar to `strptime()`, though there are some differences. To see `lubridate`'s formatting, type `?parse_date_time` into the R Console.

***
## M-P

[`magrittr`](https://CRAN.R-project.org/package=magrittr): Forward-pipe operator for R.

[`methods`](https://www.rdocumentation.org/packages/methods/versions/3.5.1): Formal methods and classes.

[`mgcv`](https://CRAN.R-project.org/package=mgcv): Mixed GAM computational vehicle with automatic smoothness estimation.

[`microbenchmark`](https://CRAN.R-project.org/package=microbenchmark): Measure and compare the execution time of R expressions.

[`odbc`](https://CRAN.R-project.org/package=odbc): Connect to ODBC compatible databases using the DBI Interface.

['openintro'](https://CRAN.R-project.org/package=openintro): Data sets and supplemental functions from OpenIntro textbooks.

[`packrat`](https://CRAN.R-project.org/package=packrat): Manage and document the versions of packages used in an R program.

* `packrat` tends to cause more trouble than it prevents, so avoid using it unless necessary or until it is improved.

[`pryr`](https://CRAN.R-project.org/package=pryr): Tools to pry back the covers of R and understand the language at a deeper level.

[`purrr`](https://CRAN.R-project.org/package=purrr): Functional programming tools.

* See also [purr.tidyverse.org](https://purrr.tidyverse.org/).

***
## Q-T

[`qdap`](https://CRAN.R-project.org/package=qdap): Bridging the gap between qualitative data and quantitative analysis (text mining).

[`qdapDictionaries`](https://CRAN.R-project.org/package=qdapDictionaries): Dictionaries and word lists for the `qdap` package.

[`RColorBrewer`](https://CRAN.R-project.org/package=RColorBrewer): ColorBrewer palattes.

[`readr`](https://CRAN.R-project.org/package=readr): Read rectangular text data.

* See also [readr.tidyverse.org](https://readr.tidyverse.org/).

[`readxl`](https://CRAN.R-project.org/package=readxl): Read Excel files.

[`rjson`](https://CRAN.R-project.org/package=rjson): Convert between R and JSON objects.

[`rlang`](https://CRAN.R-project.org/package=rlang): Functions for base types and Core R and Tidyverse features.

[`RMarkdown`](https://CRAN.R-project.org/package=rmarkdown): Save and execute code; generate high quality reports.

  * See also ["R Markdown: The Definitive Guide"](https://bookdown.org/yihui/rmarkdown/)).

[`scales`](https://CRAN.R-project.org/package=scales): Scale functions for visualization.

[`shiny`](https://CRAN.R-project.org/package=shiny): Web application framework.

[`splitstackshape`]( https://CRAN.R-project.org/package=splitstackshape): Stack and reshape datasets after splitting concatenated values.

[`stats`](https://www.rdocumentation.org/packages/stats/versions/3.5.1): Statistical functions.

[`stringr`](https://CRAN.R-project.org/package=stringr): Working with strings.

* See also [stringr.tidyverse.org](https://stringr.tidyverse.org/)).

[`svDialogs`](https://CRAN.R-project.org/package=svDialogs): Dialog boxes for Windows, MacOS, and Linuxes.

[`sys`](https://CRAN.R-project.org/package=sys): Powerful and reliable tools for running system commands in R.

[`tcltk`](https://www.rdocumentation.org/packages/tcltk/versions/3.5.2): Interface and languate bindings to Tcl/Tk GUI elements.

[`tibble`](https://CRAN.R-project.org/package=tibble): Simple data frames with stricter checking and better formatting than the traditional data frame.

[`tidyr`](https://CRAN.R-project.org/package=tidyr): Tidy data.

* See also [tidyr.tidyverse.org](https://tidyr.tidyverse.org/).

[`tinytex`](https://CRAN.R-project.org/package=tinytex): Compile LaTeX Documents.
  + Required to compile and build a `bookdown` book.
  + See https://yihui.name/tinytex/

***
## U-Z

[`utils`](https://CRAN.R-project.org/package=R.utils): Various programming utilities.

[`XLConnect`](https://CRAN.R-project.org/package=XLConnect): Read, write, and format Excel data.

[`xts`](https://CRAN.R-project.org/package=xts): Provide for uniform handling of R's different time-based data classes by extending zoo.

[`zeallot`](https://CRAN.R-project.org/package=zeallot): Multiple, unpacking, and destructuring assignment.

[`zoo`](https://CRAN.R-project.org/package=zoo): For regular and irregular time series.

<!--chapter:end:08-referenced-packages.Rmd-->

# References & Resources

* For an introduction to the R programming language, see the R Project for Statistical Computing's ["What is R?"](https://www.r-project.org/about.html) and Wikipedia's ["R (programming language)."](https://en.wikipedia.org/wiki/R_(programming_language))

* To download R, go to [r-project.org](https://www.r-project.org/) and choose the cloud CRAN Mirror option.

* To program in the R language on a user-friendly platform, download the [RStudio](https://www.rstudio.com/) IDE.

* [The R Project for Statistical Computing](https://www.r-project.org/)
  * [Library of R Packages](https://cran.r-project.org/web/packages/)
  * [*Getting Help with R*](https://www.r-project.org/help.html)
  * [*The R Manuals*](https://cran.r-project.org/manuals.html)
  * [*Frequently Asked Questions*](https://cran.r-project.org/faqs.html)
  * [*Books Related to R*](https://www.r-project.org/doc/bib/R-books.html)
  * [*Documentation*](https://www.r-project.org/other-docs.html)

* [RStudio](https://www.rstudio.com/)
  * [*RStudio Cheat Sheets*](https://www.rstudio.com/resources/cheatsheets/)
  * [*Webinars and Videos On Demand*](https://www.rstudio.com/resources/webinars/)
  * [*Online learning*](https://www.rstudio.com/online-learning/)
  * [*RStudio Blog*](https://blog.rstudio.com/)

* Online Manuals
  * [*R for Data Science*](http://r4ds.had.co.nz/)
  * [*Advanced R*](http://adv-r.had.co.nz/) by Hadley Wickham
        + Hadley's second edition draft is available [here](https://adv-r.hadley.nz/).
  * [*R Packages*](http://r-pkgs.had.co.nz/) by Hadley Wickham
  * [*The R Inferno*](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf) by Patrick Burns
  * [*The tidyverse style guide*](http://style.tidyverse.org/)
  * [*Efficient R Programming*](https://bookdown.org/csgillespie/efficientR/)
  * [*Free R Reading Material*](https://committedtotape.shinyapps.io/freeR/)

* Other Online Resources
  * [DataCamp](https://www.datacamp.com/)
  * [RDocumentation](https://www.rdocumentation.org/)
  * [R Bloggers](https://www.r-bloggers.com/how-to-learn-r-2/)
    * ["Tutorials for learning R"](https://www.r-bloggers.com/how-to-learn-r-2/)
  * [Regular Expressions 101](https://regex101.com/)

<!--chapter:end:09-references-and-resources.Rmd-->

`r if (knitr::is_html_output()) '
# References {-}
'`

<!--chapter:end:10-references.Rmd-->

